---
title: "Case study: zebrafish lethal thermal death time across life stages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: zebrafish lethal thermal death time across life stages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
```

This article fits a lethal thermal-death-time (TDT) model to zebrafish
(*Danio rerio*) measured at three points in early development, and asks a
question that ordinary thermal-limit summaries cannot: **do the life stages
differ only in how hot they tolerate (`CTmax`, `z`), or also in the *shape* of
their survival response?** It is the `freqTLS` showcase for the **shape
covariate** — letting the curve's asymptotes and steepness vary by group, not
just its thermal location.

The fit runs live by maximum likelihood (TMB, **no Stan**); uncertainty is
summarised by profile-likelihood **confidence intervals**; and the
stage-to-stage differences are **frequentist contrasts**, not posteriors.

**This vignette builds without Stan.** The `freqTLS` fits run live and need
nothing beyond this package. The Bayesian (`bayesTLS`) and classical two-stage
columns of the three-way benchmark are read from a maintainer-built cache **if it
is present**; when the cache is absent (a fresh checkout, or continuous
integration) the page shows the recipe and the `freqTLS` side and explains how
to populate it. The live `bayesTLS` (Stan) calls are documented for
reproducibility but are never evaluated here.

```{r setup}
library(freqTLS)
```

## The dataset and the question

`zebrafish_lethal` is vendored from
[`bayesTLS`](https://github.com/daniel1noble/bayesTLS) under CC BY 4.0; cite
`bayesTLS` and `citation("freqTLS")` when you use it (see `?zebrafish_lethal`).
It records survival counts in a temperature × duration grid, replicated across
three early life stages.

```{r data-shape}
data(zebrafish_lethal)
str(zebrafish_lethal)
table(zebrafish_lethal$life_stage)
```

There are `r nrow(zebrafish_lethal)` cells across the three life stages —
`r sum(zebrafish_lethal$life_stage == "young_embryos")` for young embryos,
`r sum(zebrafish_lethal$life_stage == "old_embryos")` for old embryos, and
`r sum(zebrafish_lethal$life_stage == "larvae")` for larvae — totalling
`r format(sum(zebrafish_lethal$total), big.mark = ",")` fish. The grouping factor
`life_stage` is the covariate of interest.

The biological question has two parts:

1. **Location.** Do the stages differ in thermal tolerance — the critical
   temperature `CTmax` and the thermal-sensitivity slope `z`?
2. **Shape.** Do they also differ in the *form* of the survival curve — in
   particular the maximum survival at benign exposures (the upper asymptote `up`)
   or the steepness `k`?

The first question is the standard TDT comparison. The second is where the shape
covariate earns its keep, and where the constant-shape benchmark below cannot
follow.

**Configuration.** Throughout we use the *benchmark* configuration that makes the
three estimators directly comparable (`docs/design/06-benchmark-protocol.md`):
reference time `tref = 1` hour, the **relative** mortality threshold, the
**beta-binomial** family (overdispersed survival counts), and — for the
three-way comparison — a **constant shape** (`low`, `up`, `k` shared, the
temperature effect running through the midpoint only). The shape covariate is a
deliberate, separately fitted *relaxation* of that last constraint, taken up in
the second half of this article.

## The grouped fit (live)

A single `fit_tls()` call fits all three stages at once, with stage-specific
`CTmax` and `z` but a shared constant shape — the matched benchmark
configuration.

```{r profile-zebrafish}
data(zebrafish_lethal)
zebra_fit <- suppressWarnings(fit_tls(
  zebrafish_lethal,
  y = survived, n = total, time = duration, temp = temp, group = life_stage,
  family = "beta_binomial", tref = 1
))
# per-stage CTmax with profile intervals
get_ctmax(zebra_fit)[, c("parameter", "estimate", "conf.low", "conf.high")]
```

(The zebrafish fit emits a data-adequacy warning about temperatures with fewer
than three durations; this is the kind of identifiability signal `freqTLS`
surfaces explicitly. It is suppressed here only to keep the vignette output tidy.)

The matching thermal-sensitivity slopes, again with profile-likelihood
confidence intervals:

```{r profile-zebrafish-z}
get_z(zebra_fit)[, c("parameter", "estimate", "conf.low", "conf.high")]
```

Read off the location story: **old embryos have a markedly higher `CTmax`**
(around `r round(get_ctmax(zebra_fit)$estimate[2], 1)` °C) than young embryos and
larvae (both near `r round(get_ctmax(zebra_fit)$estimate[1], 1)` °C), while the
`z` slopes are similar (roughly 1.8–2.0 °C). The profile intervals are
asymmetric where the likelihood is, and prior-free.

### A Confidence Eye per stage

The default `freqTLS` uncertainty visual is the **Confidence Eye** — a
confidence lens with a hollow point estimate, carrying no prior and making no
probability statement about the parameter. One row per stage makes the `CTmax`
separation legible at a glance:

```{r eye-ctmax, fig.alt = "Confidence Eyes for zebrafish CTmax by life stage: three pale confidence lenses with hollow point estimates, the old-embryo lens sitting clearly to the right (hotter) of the young-embryo and larval lenses; a freqTLS uncertainty display, not a posterior density."}
plot_confidence_eye(
  zebra_fit,
  parm = c("CTmax:young_embryos", "CTmax:old_embryos", "CTmax:larvae"),
  method = "profile"
)
```

The same display for the slope `z`, where the three stages overlap far more:

```{r eye-z, fig.alt = "Confidence Eyes for zebrafish z (thermal sensitivity) by life stage: three overlapping pale confidence lenses with hollow point estimates near 1.8 to 2.0 degrees Celsius, indicating similar thermal-sensitivity slopes across stages; a freqTLS uncertainty display, not a posterior density."}
plot_confidence_eye(
  zebra_fit,
  parm = c("z:young_embryos", "z:old_embryos", "z:larvae"),
  method = "profile"
)
```

### Fitted survival curves

The fitted survival curves against exposure duration, one panel per stage, with
the observed survival proportions overlaid:

```{r curves, fig.width = 7, fig.height = 4, fig.alt = "Fitted zebrafish survival curves versus exposure duration on a log time axis, faceted by life stage (young embryos, old embryos, larvae), one curve per temperature with observed survival proportions overlaid as points; survival falls with longer and hotter exposures, and the young-embryo panel plateaus at a visibly lower ceiling."}
plot_survival_curves(zebra_fit)
```

The young-embryo panel already hints at the shape story: even at benign
exposures its survival does not reach the ceiling the older stages approach. The
constant-shape fit above cannot represent that — it forces a single upper
asymptote on all three stages. We return to it after the benchmark.

## The three-way comparison, per stage

The benchmark compares three estimators of the same constant-shape 4PL, all
locked to the **relative** mortality threshold and a matched time unit /
reference time so the comparison is fair:

| Estimator | Path | Uncertainty |
| --- | --- | --- |
| Classical two-stage | `bayesTLS::ts_stage1 -> ts_stage2 -> ts_ci` | delta-method CI |
| Bayesian | `bayesTLS::fit_4pl(temp_effects = "mid") -> extract_tdt(target_surv = "relative")` | posterior credible interval |
| `freqTLS` | `fit_tls() -> confint(method = "profile")` | profile-likelihood confidence interval |

Under this matched configuration the Bayesian and profile-likelihood fits target
the **same likelihood and the same fitted curve** (see `vignette("model-math")`);
they differ only in how they summarise uncertainty — a posterior (with priors)
versus a prior-free likelihood interval.

The `bayesTLS` and two-stage numbers come from a maintainer-built cache; the
`freqTLS` column is computed live from `zebra_fit` as this page renders. The
recipe that builds the cache (shown with `eval = FALSE`, so this vignette never
needs Stan) is:

```{r bayes-recipe, eval = FALSE}
library(bayesTLS)
data(zebrafish_lethal, package = "freqTLS")

# 1. Standardise: name the temperature / duration / count columns and the time
#    unit, carrying the life_stage grouping factor through.
std <- bayesTLS::standardize_data(
  data          = zebrafish_lethal,
  temp          = "temp",
  duration      = "duration",
  n_total       = "total",
  n_surv        = "survived",
  duration_unit = "hours"
)

# 2. Bayesian fit, matched: constant shape via temp_effects = "mid",
#    beta-binomial, grouped by life stage.
bfit <- bayesTLS::fit_4pl(
  data         = std,
  group        = "life_stage",
  temp_effects = "mid",
  family       = brms::brmsfamily("beta_binomial", link = "identity"),
  chains       = 4, iter = 4000, seed = 123, backend = "cmdstanr"
)

# 3. Relative-threshold CTmax + z at tref = 1 hour, per stage.
btdt  <- bayesTLS::extract_tdt(
  bfit, target_surv = "relative",
  t_ref = 1, time_multiplier = 1, output_time_unit = "hours"
)
ctmax <- bayesTLS::get_ctmax_summary(btdt)  # per-stage temp_lower / median / upper
zsumm <- bayesTLS::get_z_summary(btdt)      # per-stage z_median / lower / upper
```

A maintainer runs `data-raw/build_benchmark_cache.R` on a Stan machine to write
`inst/extdata/bayesTLS_benchmark_cache.rds` (the Bayesian and two-stage summaries
plus a `meta` provenance block). When the cache holds the zebrafish stages, the
three-way table below fills in; otherwise it shows the live `freqTLS` column
and a note.

```{r three-way-zebra, echo = FALSE}
cache_path <- system.file("extdata", "bayesTLS_benchmark_cache.rds", package = "freqTLS")
has_cache  <- nzchar(cache_path) && file.exists(cache_path)

fmt <- function(est, lo, hi) {
  ifelse(is.na(est), "--", sprintf("%.2f [%.2f, %.2f]", est, lo, hi))
}

stage_levels <- zebra_fit$group_levels
stage_label  <- c(young_embryos = "Young embryos", old_embryos = "Old embryos",
                  larvae = "Larvae")

# Live freqTLS per-stage CTmax and z as profile-likelihood confidence
# intervals (always available; no Stan). One confint() call covers every stage;
# these are the same per-stage profiles drawn as Confidence Eyes above.
pro <- suppressMessages(confint(
  zebra_fit,
  c(paste0("CTmax:", stage_levels), paste0("z:", stage_levels)),
  method = "profile"
))
pr_of <- function(parm, g) {
  r <- pro[pro$parameter == paste0(parm, ":", g), , drop = FALSE]
  if (nrow(r) == 0L) return(stats::setNames(rep(NA_real_, 3), c("est", "lo", "hi")))
  stats::setNames(c(r$estimate[1L], r$conf.low[1L], r$conf.high[1L]),
                  c("est", "lo", "hi"))
}

# Does the cache carry the zebrafish stages? Each stage is keyed as
# "zebrafish:<life_stage>" (the stage lives in the dataset key, not a separate
# `group` column), so look for those keys rather than a bare "zebrafish".
cache_has_zebra <- FALSE
if (has_cache) {
  cache <- readRDS(cache_path)
  cache_has_zebra <- !is.null(cache$bayesian) &&
    "dataset" %in% names(cache$bayesian) &&
    all(paste0("zebrafish:", stage_levels) %in% cache$bayesian$dataset)
}

if (cache_has_zebra) {
  # Pick a cached row by stage-keyed dataset + parameter.
  pick <- function(df, parm, g, cols) {
    r <- df[df$dataset == paste0("zebrafish:", g) & df$parameter == parm, ,
            drop = FALSE]
    if (nrow(r) == 0L) return(stats::setNames(rep(NA_real_, 3), c("est", "lo", "hi")))
    stats::setNames(as.numeric(r[1L, cols]), c("est", "lo", "hi"))
  }
  one_row <- function(g, parm, quantity) {
    ts <- pick(cache$two_stage, parm, g, c("estimate", "lower", "upper"))
    by <- pick(cache$bayesian,  parm, g, c("median",   "lower", "upper"))
    pr <- pr_of(parm, g)
    data.frame(
      Stage    = stage_label[[g]],
      Quantity = quantity,
      `Two-stage (delta CI)`    = fmt(ts["est"], ts["lo"], ts["hi"]),
      `bayesTLS (95% CrI)`      = fmt(by["est"], by["lo"], by["hi"]),
      `freqTLS (profile CI)` = fmt(pr["est"], pr["lo"], pr["hi"]),
      check.names = FALSE
    )
  }
  rows <- do.call(rbind, lapply(stage_levels, function(g) rbind(
    one_row(g, "CTmax", "CTmax (°C)"),
    one_row(g, "z",     "z (°C / decade)")
  )))
  knitr::kable(rows, row.names = FALSE,
               caption = "Zebrafish CTmax and z per stage from three estimators (matched constant-shape configuration).")
} else {
  prof_tbl <- do.call(rbind, lapply(stage_levels, function(g) {
    ct <- pr_of("CTmax", g); z <- pr_of("z", g)
    data.frame(
      Stage = stage_label[[g]],
      `CTmax (°C), profile CI`        = fmt(ct["est"], ct["lo"], ct["hi"]),
      `z (°C / decade), profile CI`   = fmt(z["est"], z["lo"], z["hi"]),
      check.names = FALSE
    )
  }))
  knitr::kable(prof_tbl, row.names = FALSE,
               caption = "Live freqTLS per-stage estimates (no Stan). The bayesTLS and two-stage columns appear once the maintainer-built cache carries the zebrafish stages.")
  cat("\nNo zebrafish entries in the benchmark cache; run",
      "data-raw/build_benchmark_cache.R on a Stan machine to populate them.",
      "The freqTLS estimates above are live and need no Stan.\n")
}
```

Under the matched configuration the `freqTLS` profile-likelihood confidence
intervals and the `bayesTLS` credible intervals summarise the *same* fitted curve
per stage — one prior-free and by optimisation, the other with a prior and MCMC.
That equivalence is the point of the complementary framing; it is also exactly
what makes the benchmark a fair three-way comparison.

## Beyond the matched shape: stage-specific curves

The three-way comparison holds the shape constant (`low`, `up`, `k` shared across
temperatures and stages) so all three estimators target the *same* curve — that
is what makes the benchmark fair. But it also means the benchmark **cannot answer
the second question**: whether the stages differ in the *form* of their survival
response, not only its thermal location.

`freqTLS` can relax that restriction. Since v0.2 the shape parameters `low`,
`up`, and `log_k` may vary by a grouping factor through the `tls_bf()` formula
interface. Re-fitting `zebrafish_lethal` with fully stage-specific shapes and
comparing by AIC asks whether the life stages differ in more than thermal
*location*:

```{r v02-stage-shape}
stage_shape <- suppressWarnings(fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         CTmax ~ life_stage, log_z ~ life_stage,
         low ~ life_stage, up ~ life_stage, log_k ~ life_stage),
  data = zebrafish_lethal, family = "beta_binomial", tref = 1
))
# zebra_fit (above) is the shared-shape fit; stage_shape lets low / up / k vary.
c(shared_shape_AIC = round(AIC(zebra_fit), 1),
  stage_shape_AIC  = round(AIC(stage_shape), 1))
```

The stage-specific model has the substantially lower AIC, so the data support
per-stage shapes. The difference is concentrated in the upper asymptote `up` (the
maximum survival at benign exposures), not the steepness `k`:

```{r v02-up}
stage_est <- stage_shape$estimates
stage_est[grepl("^up:", stage_est$parameter), c("parameter", "estimate", "std.error")]
```

**This is the headline.** Young embryos have a markedly lower survival ceiling
(`up` near 0.72) than older embryos and larvae (near 0.92–0.94) — even under the
gentlest exposures, a fraction of young embryos do not survive. The stages
therefore differ not just in *where* their thermal limit sits (`CTmax`) but in
*how high their survival can ever be* (`up`). The matched constant-shape
configuration — the one the classical two-stage workflow and the `bayesTLS`
benchmark use here — forces a single ceiling on all three stages and so **cannot
express this difference at all**. The shape covariate is what makes it visible.

### Stage contrasts as frequentist intervals

The Confidence Eyes above let you *see* that the old-embryo `CTmax` sits apart.
To put a number and a confidence interval on the *difference* itself,
`freqTLS` profiles the contrast directly: it recodes the design so the
stage-to-stage difference is its own coordinate, then takes a profile-likelihood
interval on it. These are **frequentist contrasts** — differences in fitted
parameters with confidence intervals — not posterior contrasts.

```{r contrasts}
ctmax_contrasts <- rbind(
  suppressWarnings(confint(zebra_fit, "dCTmax:old_embryos-young_embryos", method = "profile")),
  suppressWarnings(confint(zebra_fit, "dCTmax:larvae-young_embryos",      method = "profile"))
)
z_contrasts <- rbind(
  suppressWarnings(confint(zebra_fit, "dlog_z:old_embryos-young_embryos", method = "profile")),
  suppressWarnings(confint(zebra_fit, "dlog_z:larvae-young_embryos",      method = "profile"))
)
rbind(ctmax_contrasts, z_contrasts)[, c("parameter", "estimate", "conf.low", "conf.high", "method")]
```

The `CTmax` contrasts confirm what the eyes show: the **old-embryo versus
young-embryo difference is about 1.46 °C with a confidence interval that
excludes zero**, while the **larva versus young-embryo difference (~0.13 °C)
straddles zero** — larvae and young embryos are indistinguishable in `CTmax`. The
`z` (slope) contrasts both straddle zero on the `log_z` scale, so there is no
compelling evidence that the stages differ in thermal sensitivity. The
stage-to-stage signal is in the *location of the limit* (old embryos elevated)
and in the *survival ceiling* (young embryos depressed), not in the slope.

(The contrast is encoded relative to the first-named stage, so the printed sign
follows that coding; the magnitude and whether the interval covers zero are the
inferential content. Contrast profiling is a fixed-effects operation; under a
random-effects fit `freqTLS` directs you to `confint(method = "wald")`
instead.)

## See also: bayesTLS

This article is the `freqTLS` half of a complementary pair. Reach for
`freqTLS` when you want fast, prior-free, asymmetry-respecting confidence
intervals; explicit identifiability checks; and — as here — a **shape covariate**
that lets the survival ceiling and steepness vary by group, with frequentist
contrasts on the differences. Reach for
[`bayesTLS`](https://github.com/daniel1noble/bayesTLS) when you want a full
Bayesian workflow, prior information, posterior contrasts, or the heat-injury and
repair sub-models. The two are lenses on the same model:
`vignette("comparing-to-bayesTLS")` lays out the relationship and the credit, and
`vignette("profile-likelihood")` covers the non-closing behaviour and the
bootstrap fallback.
