---
title: "Case 4: Drosophila suzukii mortality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case 4: Drosophila suzukii mortality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

This article follows the mortality arm of Case 4 in the
[`bayesTLS` supplement](https://daniel1noble.github.io/bayesTLS/), using the
same individual records, cell aggregation, beta-binomial response, four-hour
reference time, and absolute-LT50 comparison.

```{r setup}
library(freqTLS)
data(dsuzukii)
```

## Exact mortality aggregation

```{r data}
mort <- stats::aggregate(
  cbind(n_dead = as.integer(dsuzukii$dead),
        n_total = rep.int(1L, nrow(dsuzukii))) ~ temp + time + sex,
  data = dsuzukii,
  FUN = sum
)
mort$n_surv <- mort$n_total - mort$n_dead
mort_std <- standardize_data(
  mort, temp = "temp", duration = "time",
  n_total = "n_total", n_surv = "n_surv", duration_unit = "minutes"
)
```

## Separate-sex fits

Separate fits reproduce the study-facing comparison before the joint model.
They use the same formula components as the joint fit with one sex at a time.

```{r separate}
sex_fits <- lapply(split(mort, mort$sex), function(d) {
  d_std <- standardize_data(
    d, temp = "temp", duration = "time",
    n_total = "n_total", n_surv = "n_surv", duration_unit = "minutes"
  )
  fit_4pl(
    d_std, ctmax = ~ 1, z = ~ 1,
    low = ~ temp_c, up = ~ temp_c, k = ~ temp_c,
    family = "beta_binomial", t_ref = 240,
    method = "wald", quiet = TRUE
  )
})
lapply(sex_fits, diagnose_tdt_fit)
separate_table <- do.call(rbind, lapply(names(sex_fits), function(sex_level) {
  ans <- tls(sex_fits[[sex_level]], lethal = FALSE, method = "wald")$summary
  ans$sex <- sex_level
  names(ans)[names(ans) == "median"] <- "estimate"
  ans[c("sex", "quantity", "estimate", "lower", "upper")]
}))
separate_table
```

## Joint direct model

The direct joint model assigns separate `CTmax` and `z` coordinates to each sex
and places centred temperature on `low`, `up`, and `k`. The shape coefficients
are experimental in freqTLS. The very large standard errors that can arise for
the shape slopes are an identifiability result and must remain visible.

```{r joint}
mort_fit <- fit_4pl(
  mort_std,
  ctmax = ~ 0 + sex,
  z = ~ 0 + sex,
  low = ~ temp_c,
  up = ~ temp_c,
  k = ~ temp_c,
  family = "beta_binomial",
  t_ref = 240,
  method = "wald",
  quiet = TRUE
)
diagnose_tdt_fit(mort_fit)
check_tls(mort_fit)
```

The optimizer and Hessian pass, but the lower-asymptote coefficients remain
weakly identified. Their actual estimates and standard errors are shown rather
than hidden behind the successful convergence summary.

```{r shape-identification}
mort_shape <- subset(
  mort_fit$fit$estimates,
  grepl("^(low|up|k):", parameter),
  select = c(parameter, estimate, std.error)
)
mort_shape
```

```{r quantities}
mort_relative <- tls(mort_fit, by = "sex", lethal = FALSE, method = "wald")$summary
names(mort_relative)[names(mort_relative) == "median"] <- "estimate"
mort_relative
```

The study-facing threshold is absolute LT50. The fitted backbone uses the
relative coordinate. Because the shape also changes with assay temperature,
the absolute 50% point is obtained by solving the fitted survival curve at 240
minutes for each sex.

```{r absolute}
mort_lt50 <- data.frame(
  sex = mort_fit$fit$group_levels,
  LT50_240_min = vapply(mort_fit$fit$group_levels, function(sex_level) {
    objective <- function(temp) {
      predict(
        mort_fit,
        data.frame(temp = temp, duration = 240, group = sex_level),
        type = "survival"
      ) - 0.5
    }
    uniroot(objective, range(mort$temp), tol = 1e-10)$root
  }, numeric(1)),
  interval_method = "none: exact-model bootstrap unstable"
)
mort_lt50
```

This is an ML point comparison, not a confidence interval. The exact-model
parametric bootstrap produced too few converged refits, so freqTLS does not
print a misleading interval here. Use the pinned bayesTLS Case 4 posterior fit
for the independently regularized absolute-LT50 interval.

```{r fitted-curves, fig.alt="Observed survival proportions and fitted Drosophila suzukii mortality curves, separated by sex and assay temperature."}
plot_survival_curves(mort_fit$fit)
```

The pinned bayesTLS comparison uses its recorded specification; current bayesTLS
also supports direct `CTmax`/`z` parameterisation. freqTLS keeps direct `CTmax`/
`z` parameters so likelihood profiles can target them directly. It does not
report `Tcrit` as a Case 4 result.
