---
title: "Case 2: cereal-aphid thermal tolerance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case 2: cereal-aphid thermal tolerance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

This article follows the cereal-aphid case in the
[`bayesTLS` supplement](https://daniel1noble.github.io/bayesTLS/): the heat
branch, 6-day-old aphids for the main analysis, then the all-age heat extension.

```{r setup}
library(freqTLS)
data(aphid_tdt)
```

## Main analysis: heat branch at age 6

```{r data}
aphid6 <- droplevels(subset(aphid_tdt, branch == "heat" & age == "6"))
aphid6$species <- factor(
  aphid6$species,
  levels = c("M_dirhodum", "R_padi", "S_avenae"),
  labels = c("Metopolophium dirhodum", "Rhopalosiphum padi",
             "Sitobion avenae")
)
aphid6_std <- standardize_data(
  aphid6, temp = "temp", duration = "duration_min",
  n_total = "n_total", n_surv = "n_surv", duration_unit = "minutes"
)
```

The beta-binomial model gives each species a direct `CTmax` and `z`, lets
steepness vary with centred assay temperature, and shares `low` and `up`.
`t_ref = 60` is one hour because the standardized duration unit is minutes.

```{r fit}
aphid6_fit <- fit_4pl(
  aphid6_std,
  ctmax = ~ 0 + species,
  z = ~ 0 + species,
  low = ~ 1,
  up = ~ 1,
  k = ~ temp_c,
  family = "beta_binomial",
  t_ref = 60,
  method = "wald",
  quiet = TRUE
)
diagnose_tdt_fit(aphid6_fit)
```

```{r quantities}
aphid6_tls <- tls(aphid6_fit, by = "species", lethal = FALSE, method = "wald")
aphid6_table <- aphid6_tls$summary
names(aphid6_table)[names(aphid6_table) == "median"] <- "estimate"
aphid6_table
```

This experimental freqTLS page reports `CTmax` and `z` only. The pinned
bayesTLS case additionally extracts `Tcrit`; users needing that Bayesian
derived quantity should use the supplement rather than interpret its omission
here as a zero biological effect.

```{r tdt, fig.alt="Thermal death-time curves for three cereal aphid species in the heat branch at age six."}
plot_tdt_curve(aphid6_fit$fit)
```

The table reports Wald confidence intervals. The TDT figure shows fitted point
curves, not uncertainty intervals. Check convergence, Hessian status, gradient,
and data-adequacy warnings before ranking species.

## All-age heat extension

The supplement next uses all heat-branch ages and the direct fixed structure
`species * age` on both headline coordinates. This is an experimental larger
fixed-effect model in freqTLS; it is not a substitute empirical case.

```{r all-age}
aphid_all <- droplevels(subset(aphid_tdt, branch == "heat"))
aphid_all$species <- factor(
  aphid_all$species,
  levels = c("M_dirhodum", "R_padi", "S_avenae"),
  labels = c("Metopolophium dirhodum", "Rhopalosiphum padi",
             "Sitobion avenae")
)
aphid_all_std <- standardize_data(
  aphid_all, temp = "temp", duration = "duration_min",
  n_total = "n_total", n_surv = "n_surv", duration_unit = "minutes"
)
aphid_all_fit <- fit_4pl(
  aphid_all_std,
  ctmax = ~ 1 + species * age,
  z = ~ 1 + species * age,
  low = ~ 1,
  up = ~ 1,
  k = ~ temp_c,
  family = "beta_binomial",
  t_ref = 60,
  method = "wald",
  quiet = TRUE
)
diagnose_tdt_fit(aphid_all_fit)
```

The interacted formula coefficients are not themselves species-by-age values.
The parameter prediction below evaluates the fitted direct coordinates for all
nine biological cells at the observed mean assay temperature.

```{r all-age-quantities}
aphid_cells <- expand.grid(
  species = levels(aphid_all$species),
  age = levels(aphid_all$age),
  temp = mean(aphid_all_std$temp),
  KEEP.OUT.ATTRS = FALSE
)
aphid_cell_pars <- predict(aphid_all_fit, aphid_cells, type = "parameters")
cbind(aphid_cells[c("species", "age")],
      aphid_cell_pars[c("CTmax", "z")])
```

```{r all-age-tdt, fig.alt="Thermal death-time curves for each cereal aphid species, faceted by age, from the all-age heat-branch model."}
aphid_tdt_grid <- merge(
  expand.grid(
    temp = seq(min(aphid_all_std$temp), max(aphid_all_std$temp), length.out = 60),
    KEEP.OUT.ATTRS = FALSE
  ),
  aphid_cells[c("species", "age")]
)
aphid_tdt_pars <- predict(aphid_all_fit, aphid_tdt_grid, type = "parameters")
aphid_tdt_grid$duration_50 <- 10^(
  log10(60) - (aphid_tdt_grid$temp - aphid_tdt_pars$CTmax) / aphid_tdt_pars$z
)
ggplot2::ggplot(
  aphid_tdt_grid,
  ggplot2::aes(temp, duration_50, colour = species)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_log10() +
  ggplot2::facet_wrap(~ age) +
  ggplot2::labs(
    x = "Temperature (°C)",
    y = "Duration to 50% relative survival (minutes, log scale)",
    colour = "Species"
  ) +
  ggplot2::theme_minimal()
```

## Projection boundary

The Wuhan field-temperature projection is not run here. Its environmental
trace remains outside the installed package until the full redistribution chain
is recorded in the licence ledger. `predict_heat_injury()` can evaluate a
temperature trace, but any repair kernel supplied to current freqTLS is
illustrative and unfitted. The fitted Bayesian projection remains available in
bayesTLS.
