---
title: "Case 3: Snow-gum PSII under Dark and Light recovery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case 3: Snow-gum PSII under Dark and Light recovery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

This is the frequentist analogue of the Snow-gum PSII case in the
[`bayesTLS` supplement](https://daniel1noble.github.io/bayesTLS/). The response
is retained photosystem-II function, not mortality, so the experimental Beta
family is used.

```{r setup}
library(freqTLS)
data(snowgum_psii)
```

## Data and boundary handling

`fvfm_prop` is post-exposure Fv/Fm divided by its pre-exposure value. The
394-row object is byte-identical to the pinned bayesTLS object. Exact zeros are
clamped inward by `standardize_data()` because a Beta density is defined on
the open interval `(0, 1)`; the warning makes this adjustment visible.

```{r data}
leaf_std <- standardize_data(
  snowgum_psii,
  temp = "Temp", duration = "Time", proportion = "fvfm_prop",
  duration_unit = "minutes"
)
table(snowgum_psii$recovery)
```

## Locked model

Recovery condition has direct effects on `CTmax` and `z`; plant supplies one
independent random intercept on `CTmax`; `low`, `up`, and `k` are shared. The
reference time is 60 minutes and the threshold is the relative midpoint.

This shared-shape choice is an intentional freqTLS analogue, not a claim of
formula identity: the pinned bayesTLS supplement also teaches
recovery-by-temperature shape terms. Those additional Bayesian shape terms are
not silently substituted here.

```{r fit}
leaf_fit <- fit_4pl(
  leaf_std,
  ctmax = ~ 0 + recovery + (1 | plant),
  z = ~ 0 + recovery,
  low = ~ 1,
  up = ~ 1,
  k = ~ 1,
  family = "beta",
  t_ref = 60,
  method = "wald",
  quiet = TRUE
)
diagnose_tdt_fit(leaf_fit)
```

```{r quantities}
leaf_tls <- tls(leaf_fit, by = "recovery", lethal = FALSE, method = "wald")
leaf_table <- leaf_tls$summary
names(leaf_table)[names(leaf_table) == "median"] <- "estimate"
leaf_table
```

```{r eye, fig.alt="Wald Confidence Eyes for Snow-gum CTmax and z after Dark and Light recovery."}
plot_confidence_eye(leaf_fit, parm = c("CTmax", "z"), method = "wald")
```

This sublethal endpoint reports only `CTmax` and `z`; `Tcrit` is not a valid
headline quantity here. Inspect convergence, Hessian/gradient diagnostics, the
Beta boundary warning, and the limited-random-intercept assumptions before
interpretation.

## Licence boundary

Arnold et al. (2026; doi:10.64898/2026.04.09.717599) release the current source
under CC BY-NC 4.0. Pieter A. Arnold, the data holder, authorised its use as
this package's separately licensed teaching dataset. Retain the source and
licence notice when reusing it. See `inst/COPYRIGHTS` and the licence ledger.
