---
title: "Cross-case-study summary: four taxa, one panel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-case-study summary: four taxa, one panel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 6,
  dpi = 96
)
```

This vignette mirrors Manuscript Figure 5 of the `bayesTLS` supplement -- a single
multi-taxon panel of the thermal-sensitivity parameter `z` and the critical
temperature `CTmax` across all four case studies -- but draws the `freqTLS`
**Confidence Eye** for each estimate instead of a Bayesian posterior ridge. Each
row is a likelihood **confidence interval** (a pale lens with a hollow point
estimate), never a posterior density.

**This vignette builds without Stan.** Every fit here is the `freqTLS`
maximum-likelihood path (TMB), run live as the page renders. The Bayesian
multi-taxon ridge plot is the `bayesTLS` complement; see the closing note.

```{r setup}
library(freqTLS)
```

## The four taxa and the cross-study question

The four shared case-study datasets span four corners of thermal physiology:

* **brown shrimp** *Crangon crangon* -- a crustacean, lethal mortality, ungrouped;
* **zebrafish** *Danio rerio* -- a fish, lethal mortality, across three life
  stages (young embryos, old embryos, larvae);
* **snow gum** *Eucalyptus pauciflora* -- a plant, a **functional** (sublethal)
  endpoint, the retained proportion of photosystem-II efficiency (`Fv/Fm`);
* **vinegar fly** *Drosophila suzukii* -- an insect, lethal mortality, across the
  two sexes.

The cross-study question is descriptive: laid side by side, how do thermal limits
compare across a crustacean, a fish (resolved by developmental stage), a plant's
photosynthetic machinery, and an insect (resolved by sex)? `CTmax` (the critical
temperature at a fixed reference exposure) places each taxon on the temperature
axis; `z` (the change in temperature, in degrees Celsius, that multiplies tolerated
exposure time tenfold) measures how steeply tolerance trades off against duration.

**One caveat governs the whole panel.** The reference exposures differ by study,
because each follows the convention of its source assay:

| Taxon | Endpoint | Reference exposure (`tref`) | Threshold |
| --- | --- | --- | --- |
| Shrimp | lethal | 1 hour | relative midpoint |
| Zebrafish (per stage) | lethal | 1 hour | relative midpoint |
| Snow gum | sublethal PSII | 1 hour | relative midpoint |
| *D. suzukii* (per sex) | lethal | 4 hours (240 min) | absolute LT50 |

Because a `CTmax` is defined *at* its reference exposure, the four `CTmax` values
are **not** on a single common time scale: the snow gum number is "the temperature
that halves PSII over 1 hour", the fly number is "the temperature that kills
half the flies over 4 hours". The panel is therefore an **illustrative
cross-taxon synthesis**, not a single common-scale comparison. `z`, by contrast,
is a slope -- degrees per tenfold change in time -- and is comparable across taxa
regardless of the reference exposure. Read the `CTmax` facet as four study-specific
anchors and the `z` facet as a like-with-like comparison of duration sensitivity.

## The four fits (run live)

Each fit uses the configuration locked for its case study. The grouped fits
(zebrafish, fly) estimate a separate `CTmax` and `z` per level with a shared shape
(`low`, `up`, `k`); the ungrouped fits (shrimp, snow gum) estimate one of each. The
snow gum fit uses the **beta** family for its continuous proportion; the others use
**beta-binomial** survival counts. All fits are wrapped in `suppressWarnings()` to
swallow the honest data-adequacy notes (for example, temperatures with few
durations) that `freqTLS` raises -- those signals are covered in the per-study
vignettes.

The vinegar-fly data ships per-individual (`dsuzukii`); we aggregate it to
`(temp, time, sex)` counts in base R, exactly as the per-study vignette does.

```{r fits}
# Each taxon: standardize_data() its raw table, then fit_4pl(); $fit is the
# engine fit the panel below reads.

# Shrimp: ungrouped, 1-hour reference.
data(shrimp_lethal)
shrimp_fit <- suppressWarnings(fit_4pl(standardize_data(
  shrimp_lethal, temp = "Temperature_assay", duration = "Duration_exposure_hours",
  n_total = "N_individuals_after_trial", mortality = "Mortality_after_trial",
  duration_unit = "hours"), t_ref = 1, family = "beta_binomial", quiet = TRUE))$fit

# Zebrafish: grouped by life stage, 1-hour reference.
data(zebrafish_lethal)
zebra_fit <- suppressWarnings(fit_4pl(standardize_data(
  zebrafish_lethal, temp = "assay_temp", duration = "duration_h",
  n_total = "n_total", n_surv = "n_surv", duration_unit = "hours"),
  by = "life_stage", t_ref = 1, family = "beta_binomial", quiet = TRUE))$fit

# Snow gum: ungrouped, beta family, 1-hour reference (functional PSII).
data(snowgum_psii)
snowgum_fit <- suppressWarnings(fit_4pl(standardize_data(
  snowgum_psii, temp = "Temp", duration = "Time", proportion = "fvfm_prop",
  duration_unit = "minutes"), family = "beta", t_ref = 60, quiet = TRUE))$fit

# D. suzukii: per-individual -> (temp, time, sex) counts; grouped by sex, 4-hour ref.
data(dsuzukii)
.nd <- aggregate(list(n_dead  = dsuzukii$dead), dsuzukii[c("temp", "time", "sex")], sum)
.nt <- aggregate(list(n_total = dsuzukii$dead), dsuzukii[c("temp", "time", "sex")], length)
fly_fit <- suppressWarnings(fit_4pl(standardize_data(
  merge(.nd, .nt), temp = "temp", duration = "time",
  n_total = "n_total", n_dead = "n_dead", duration_unit = "minutes"),
  by = "sex", t_ref = 240, family = "beta_binomial", quiet = TRUE))$fit
```

## Assembling the cross-study table

The panel is built from one combined data frame of
`(label, parameter, estimate, conf.low, conf.high)`, obtained by calling
`confint(..., method = "profile")` on each fit. For the grouped fits we ask for the
per-level parameter names (`CTmax:young_embryos`, `z:M`, and so on); for the
ungrouped fits we ask for the bare `CTmax` and `z`. Every interval here is a
profile-likelihood confidence interval, and -- as the `conf.status` column
confirms -- every profile closes.

```{r combine}
# A small helper: pull profile CIs for the requested parameters from one fit and
# tag each row with a human-readable taxon/group label.
eye_rows <- function(fit, taxon, parms, labels) {
  ci <- suppressWarnings(suppressMessages(
    confint(fit, parm = parms, method = "profile")
  ))
  base <- sub(":.*$", "", ci$parameter)            # "CTmax:M" -> "CTmax"
  data.frame(
    taxon     = taxon,
    label     = labels[match(ci$parameter, parms)],
    parameter = base,
    estimate  = ci$estimate,
    conf.low  = ci$conf.low,
    conf.high = ci$conf.high,
    status    = ci$conf.status,
    stringsAsFactors = FALSE
  )
}

panel <- rbind(
  eye_rows(shrimp_fit, "Shrimp (1 h)",
           c("CTmax", "z"),
           c("Shrimp", "Shrimp")),
  eye_rows(zebra_fit, "Zebrafish (1 h)",
           c("CTmax:young_embryos", "CTmax:old_embryos", "CTmax:larvae",
             "z:young_embryos", "z:old_embryos", "z:larvae"),
           c("Zebrafish: young embryos", "Zebrafish: old embryos",
             "Zebrafish: larvae", "Zebrafish: young embryos",
             "Zebrafish: old embryos", "Zebrafish: larvae")),
  eye_rows(snowgum_fit, "Snow gum (1 h)",
           c("CTmax", "z"),
           c("Snow gum (PSII)", "Snow gum (PSII)")),
  eye_rows(fly_fit, "D. suzukii (4 h)",
           c("CTmax:F", "CTmax:M", "z:F", "z:M"),
           c("D. suzukii: female", "D. suzukii: male",
             "D. suzukii: female", "D. suzukii: male"))
)

# Seven rows, two parameters: a tidy printout of the headline numbers.
panel_wide <- data.frame(
  Group           = panel$label[panel$parameter == "CTmax"],
  `CTmax estimate` = round(panel$estimate[panel$parameter == "CTmax"], 2),
  `CTmax 95% CI`  = sprintf("[%.2f, %.2f]",
                            panel$conf.low[panel$parameter == "CTmax"],
                            panel$conf.high[panel$parameter == "CTmax"]),
  `z estimate`     = round(panel$estimate[panel$parameter == "z"], 2),
  `z 95% CI`       = sprintf("[%.2f, %.2f]",
                            panel$conf.low[panel$parameter == "z"],
                            panel$conf.high[panel$parameter == "z"]),
  check.names = FALSE
)
knitr::kable(
  panel_wide,
  caption = "Seven taxon/group rows: CTmax (at the study reference exposure) and z, each with its profile-likelihood 95% confidence interval."
)
```

## Cross-taxon validation against bayesTLS and the two-stage estimator

The same headline numbers, now beside the classical two-stage estimator and the
`bayesTLS` posterior, read from the maintainer-built benchmark cache. Each fit
uses its study's matched configuration (relative-midpoint threshold, constant
shape, the per-study reference exposure), so the three estimators target the
*same* fitted curve per row. Snow-gum PSII is a continuous proportion with no
count representation, so its two-stage cell is empty (it is a `bayesTLS` beta
two-way). The `freqTLS` profile-likelihood confidence intervals and the
`bayesTLS` credible intervals land on essentially the same `CTmax` and `z` in
every row — the cross-taxon statement of the complementary framing.

```{r three-way-summary, echo = FALSE}
cache_path <- system.file("extdata", "bayesTLS_benchmark_cache.rds", package = "freqTLS")
fmt3 <- function(v) ifelse(is.na(v[1]), "--", sprintf("%.2f [%.2f, %.2f]", v[1], v[2], v[3]))

# Map each summary row label to its benchmark-cache dataset key.
keys <- c(
  "Shrimp"                   = "shrimp",
  "Zebrafish: young embryos" = "zebrafish:young_embryos",
  "Zebrafish: old embryos"   = "zebrafish:old_embryos",
  "Zebrafish: larvae"        = "zebrafish:larvae",
  "Snow gum (PSII)"          = "snowgum",
  "D. suzukii: female"       = "dsuzukii:F",
  "D. suzukii: male"         = "dsuzukii:M"
)

if (nzchar(cache_path) && file.exists(cache_path)) {
  cache <- readRDS(cache_path)
  cache_pick <- function(df, key, parm, cols) {
    if (is.null(df)) return(c(NA_real_, NA_real_, NA_real_))
    r <- df[df$dataset == key & df$parameter == parm, , drop = FALSE]
    if (nrow(r) == 0L) return(c(NA_real_, NA_real_, NA_real_))
    as.numeric(r[1L, cols])
  }
  pro_pick <- function(label, parm) {
    r <- panel[panel$label == label & panel$parameter == parm, , drop = FALSE]
    if (nrow(r) == 0L) return(c(NA_real_, NA_real_, NA_real_))
    c(r$estimate[1L], r$conf.low[1L], r$conf.high[1L])
  }
  quantities <- list(c("CTmax", "CTmax (°C)"), c("z", "z (°C / decade)"))
  rows <- do.call(rbind, lapply(names(keys), function(label) {
    key <- keys[[label]]
    do.call(rbind, lapply(quantities, function(q) {
      ts <- cache_pick(cache$two_stage, key, q[1], c("estimate", "lower", "upper"))
      by <- cache_pick(cache$bayesian,  key, q[1], c("median", "lower", "upper"))
      pr <- pro_pick(label, q[1])
      data.frame(
        Group = label, Quantity = q[2],
        `Two-stage (delta CI)`    = fmt3(ts),
        `bayesTLS (95% CrI)`      = fmt3(by),
        `freqTLS (profile CI)` = fmt3(pr),
        check.names = FALSE
      )
    }))
  }))
  knitr::kable(rows, row.names = FALSE,
               caption = "Cross-taxon three-way validation: CTmax and z per taxon/group from the classical two-stage estimator, bayesTLS (posterior median + 95% CrI), and freqTLS (profile-likelihood estimate + 95% CI), each on its study's matched configuration. freqTLS and bayesTLS agree across every row.")
} else {
  cat("Benchmark cache not found; run data-raw/build_benchmark_cache.R to populate",
      "the bayesTLS and two-stage columns.\n")
}
```

## The panel: a Confidence Eye per taxon and group

Each row is drawn as an honest Confidence Eye -- a pale, shallow horizontal lens
whose width is exactly the 95% confidence interval and whose cosine taper is
tallest at the estimate, with a hollow point marking the estimate itself. The two
facets carry independent x-axes (`CTmax` in degrees Celsius; `z` in degrees Celsius
per tenfold change in exposure time), because the two parameters live on different
scales and, for `CTmax`, on different reference exposures.

The geometry follows the `freqTLS` Confidence-Eye contract: the shallow,
wide lens reads as a confidence *interval*, not a probability density, and a
profile that did not close would render as a hollow point with **no** lens. All
seven profiles here close, so every row carries a lens.

```{r panel, fig.width = 7.5, fig.height = 6, fig.alt = "Two-facet Confidence-Eye panel. Left facet, CTmax in degrees Celsius: seven pale horizontal confidence-interval lenses with hollow point estimates, one per taxon/group, ordered top to bottom as shrimp near 31.8, the three zebrafish stages near 39.8 to 41.4, snow gum PSII near 44.6, and the two D. suzukii sexes near 35.2. Right facet, z in degrees Celsius per tenfold time change: matching lenses with shrimp near 2.2, zebrafish stages near 1.8 to 2.0, snow gum near 3.7, and the two fly sexes near 3.0 to 3.2. Each CTmax row is annotated with its reference exposure (1 hour or 4 hours), underscoring that the CTmax values are not on a common time scale."}
# Build the honest Confidence-Eye geometry by hand so all four fits share one
# cross-taxon panel: a cosine-tapered pale lens (width = confidence interval)
# plus a hollow point, faceted by parameter with free x-axes. This reuses the
# package lens shape (see ?plot_confidence_eye) but spans every fit in one figure.
stopifnot(requireNamespace("ggplot2", quietly = TRUE))

# Fixed top-to-bottom row order (shrimp, zebrafish stages, snow gum, fly sexes).
row_order <- c(
  "Shrimp",
  "Zebrafish: young embryos", "Zebrafish: old embryos", "Zebrafish: larvae",
  "Snow gum (PSII)",
  "D. suzukii: female", "D. suzukii: male"
)
panel$row <- match(panel$label, row_order)
panel$parameter <- factor(panel$parameter, levels = c("CTmax", "z"),
                          labels = c("CTmax (°C)", "z (°C / decade time)"))

# One cosine-tapered lens polygon per row (tallest at the estimate, zero at each
# bound), built per facet so the free x-axes do not distort the taper.
lens_df <- do.call(rbind, lapply(seq_len(nrow(panel)), function(i) {
  x <- seq(panel$conf.low[i], panel$conf.high[i], length.out = 80)
  d_lo <- max(panel$estimate[i] - panel$conf.low[i], .Machine$double.eps)
  d_hi <- max(panel$conf.high[i] - panel$estimate[i], .Machine$double.eps)
  frac <- ifelse(x <= panel$estimate[i], (panel$estimate[i] - x) / d_lo,
                 (x - panel$estimate[i]) / d_hi)
  w <- 0.32 * cos((pi / 2) * pmin(pmax(frac, 0), 1))
  data.frame(id = i, parameter = panel$parameter[i], x = x,
             ymin = panel$row[i] - w, ymax = panel$row[i] + w)
}))

# Reference-exposure annotation for the CTmax facet only.
tref_lab <- data.frame(
  parameter = factor("CTmax (°C)",
                     levels = levels(panel$parameter)),
  row = panel$row[panel$parameter == "CTmax (°C)"],
  x   = panel$conf.high[panel$parameter == "CTmax (°C)"],
  lab = c("1 h", "1 h", "1 h", "1 h", "1 h", "4 h", "4 h")
)

ggplot2::ggplot() +
  ggplot2::geom_ribbon(
    data = lens_df,
    ggplot2::aes(x = x, ymin = ymin, ymax = ymax, group = id),
    fill = "#1b7837", colour = NA, alpha = 0.30
  ) +
  ggplot2::geom_point(
    data = panel,
    ggplot2::aes(x = estimate, y = row),
    shape = 21, fill = "white", colour = "#1b7837", size = 3, stroke = 1
  ) +
  ggplot2::geom_text(
    data = tref_lab,
    ggplot2::aes(x = x, y = row, label = lab),
    hjust = -0.25, size = 2.7, colour = "grey35"
  ) +
  ggplot2::scale_y_continuous(
    breaks = seq_along(row_order), labels = row_order,
    trans = "reverse", expand = ggplot2::expansion(add = 0.7)
  ) +
  ggplot2::facet_wrap(~ parameter, scales = "free_x") +
  ggplot2::labs(
    x = NULL, y = NULL,
    title = "Thermal limits across four taxa: CTmax and z",
    subtitle = "Confidence Eyes: pale lens = 95% confidence interval; hollow point = estimate.",
    caption = paste(
      "Profile-likelihood confidence intervals (freqTLS, no Stan).",
      "CTmax reference exposure differs by study (annotated): not a common time scale."
    )
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    plot.caption = ggplot2::element_text(hjust = 0)
  )
```

## Within-taxon contrasts

The two grouped studies invite a within-taxon comparison: do the zebrafish life
stages differ, and do the fly sexes differ? `freqTLS` answers these with a
profile-likelihood interval **on the difference** (`dCTmax:A-B`, `dz:A-B`) -- a
prior-free, frequentist counterpart to the `bayesTLS` pMCMC bracket. A difference
interval that excludes zero is a clear separation; one that spans zero is not.

```{r contrasts}
zebra_contrasts <- suppressWarnings(suppressMessages(confint(
  zebra_fit,
  parm = c("dCTmax:old_embryos-young_embryos",
           "dCTmax:larvae-young_embryos",
           "dCTmax:larvae-old_embryos",
           "dz:old_embryos-young_embryos",
           "dz:larvae-young_embryos",
           "dz:larvae-old_embryos"),
  method = "profile"
)))

fly_contrasts <- suppressWarnings(suppressMessages(confint(
  fly_fit,
  parm = c("dCTmax:M-F", "dz:M-F"),
  method = "profile"
)))

contrast_tbl <- rbind(zebra_contrasts, fly_contrasts)
knitr::kable(
  data.frame(
    Contrast   = contrast_tbl$parameter,
    Difference = round(contrast_tbl$estimate, 3),
    `95% CI`   = sprintf("[%.3f, %.3f]",
                         contrast_tbl$conf.low, contrast_tbl$conf.high),
    `Excludes 0` = ifelse(
      contrast_tbl$conf.low > 0 | contrast_tbl$conf.high < 0, "yes", "no"
    ),
    check.names = FALSE
  ),
  caption = "Within-taxon contrasts: profile-likelihood 95% confidence intervals on the CTmax and z differences."
)
```

For zebrafish, the one clear separation in `CTmax` is **old embryos versus young
embryos** and **larvae versus old embryos**: old embryos sit about 1.46 degrees
Celsius above young embryos, with a confidence interval well clear of zero,
while larvae are essentially indistinguishable from young embryos in `CTmax`. None
of the `z` contrasts excludes zero, so the three stages share a common duration
sensitivity even where their critical temperatures differ. For *D. suzukii*, both
the `CTmax` and the `z` sex contrasts span zero: there is no clear sex difference
in either thermal limit, the same conclusion Ørsted et al. (2024) reached for these
data.

## What this shows

Three things stand out from the panel:

* **The plant's photosynthetic endpoint is the steepest.** Snow gum PSII has the
  largest `z` (about 3.7 degrees Celsius per tenfold time), so its tolerated
  exposure collapses faster with temperature than the animals' lethal endpoints
  (`z` between roughly 1.8 and 3.2). Its `CTmax` is also the highest, on the same
  one-hour reference as the shrimp and zebrafish; only the *D. suzukii* `CTmax`
  sits at a different (four-hour) reference, so that row alone is off the common
  time scale.

* **Resolving a taxon by an internal axis can matter or not.** Zebrafish life stage
  shifts `CTmax` by over a degree (old embryos are the most heat-tolerant stage),
  whereas *D. suzukii* sex shifts neither `CTmax` nor `z` detectably. The
  Confidence Eyes make this visible at a glance: the zebrafish lenses separate on
  the `CTmax` axis, while the two fly lenses overlap almost completely.

* **Every estimate carries an honest, prior-free interval.** All seven profiles
  close, so each row is a closed lens rather than a hollow point. Where a profile
  did **not** close (a weakly identified design or a boundary asymptote), the same
  display would show a hollow point with no lens -- never a fabricated closed eye.

These are confidence intervals throughout: they summarise the
likelihood's support for each parameter and make no probability statement about the
parameter itself.

## See bayesTLS for the Bayesian multi-taxon ridge

This panel is the `freqTLS` (likelihood) counterpart to Manuscript Figure 5 of
the `bayesTLS` supplement, which draws the same four taxa as posterior **ridge
densities** with median points and 95% credible bars. The two displays answer the
same cross-study question from complementary inferential engines: the `bayesTLS`
ridge is a posterior density shaped by priors and the data, whereas the
`freqTLS` Confidence Eye is a prior-free likelihood interval that cannot be
read as a posterior. For the Bayesian multi-taxon ridge plot, the within-taxon
pMCMC contrasts, and the full posterior workflow, see `bayesTLS`
(<https://github.com/daniel1noble/bayesTLS>); for the side-by-side
posterior-versus-Confidence-Eye contrast on a single dataset, see
`vignette("comparing-to-bayesTLS")`.
