---
title: "Case 4.2: Drosophila suzukii awake/coma response"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case 4.2: Drosophila suzukii awake/coma response}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

Case 4.2 in the
[`bayesTLS` supplement](https://daniel1noble.github.io/bayesTLS/) contains
several analyses. freqTLS supports only the awake/coma count arm. It does not
replace the censored time-to-coma or hurdle-productivity models with a different
endpoint.

```{r setup}
library(freqTLS)
data(dsuzukii)
```

## Exact awake-count aggregation

Within each temperature x exposure-level x sex cell, `n_awake` is the number of
individuals with missing `t_coma`. Duration is the first recorded `time` in the
cell, and duration-zero controls are dropped.

```{r data}
cell <- interaction(dsuzukii$temp, dsuzukii$lvl, dsuzukii$sex, drop = TRUE)
coma <- do.call(rbind, lapply(split(dsuzukii, cell), function(d) {
  data.frame(
    temp = d$temp[1], lvl = d$lvl[1], sex = d$sex[1],
    duration = d$time[1], n_total = nrow(d),
    n_awake = sum(is.na(d$t_coma))
  )
}))
coma <- droplevels(subset(coma, duration > 0))
coma_std <- standardize_data(
  coma, temp = "temp", duration = "duration",
  n_total = "n_total", n_surv = "n_awake", duration_unit = "minutes"
)
```

## Locked model

The awake/coma response is beta-binomial. The direct headline formulas retain
an intercept (`~ sex`), while `low`, `up`, and `k` each vary with centred assay
temperature. The reference is 60 minutes and the estimand is the relative
midpoint.

```{r fit}
coma_fit <- fit_4pl(
  coma_std,
  ctmax = ~ sex,
  z = ~ sex,
  low = ~ temp_c,
  up = ~ temp_c,
  k = ~ temp_c,
  family = "beta_binomial",
  t_ref = 60,
  method = "wald",
  quiet = TRUE
)
diagnose_tdt_fit(coma_fit)
check_tls(coma_fit)
```

The lower-asymptote coefficients are weakly identified even though the
optimizer, Hessian, and raw-gradient checks pass. The actual estimates and
standard errors make that limitation visible.

```{r shape-identification}
coma_shape <- subset(
  coma_fit$fit$estimates,
  grepl("^(low|up|k):", parameter),
  select = c(parameter, estimate, std.error)
)
coma_shape
```

```{r quantities}
coma_table <- tls(coma_fit, by = "sex", lethal = FALSE, method = "wald")$summary
names(coma_table)[names(coma_table) == "median"] <- "estimate"
coma_table
```

Only `CTmax` and `z` are reported; `Tcrit` is not. Large shape-coefficient
standard errors or warnings are part of the result, not material to hide.

```{r fitted-curves, fig.alt="Observed awake proportions and fitted Drosophila suzukii awake-coma curves, separated by sex and assay temperature."}
plot_survival_curves(coma_fit$fit)
```

## Unsupported arms

The individual `t_coma` values form a censored time-to-event outcome, and
`prod` requires a hurdle model for zero reproduction plus positive clutch size.
freqTLS 0.2.0.9000 implements neither. Use the corresponding bayesTLS Case 4.2
sections for those analyses rather than treating the count model above as a
replacement.
