---
title: "Case 1: zebrafish heat tolerance under normoxia and hyperoxia"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case 1: zebrafish heat tolerance under normoxia and hyperoxia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

This is the frequentist analogue of Case 1 in the
[`bayesTLS` supplement](https://daniel1noble.github.io/bayesTLS/). It uses the
same organism, data, filter, response, formulas, one-hour reference time,
relative threshold, and estimands. Only the inference engine and uncertainty
language differ.

```{r setup}
library(freqTLS)
```

## Exact data subset

We retain diploid larvae under normoxia and hyperoxia, including the 26 °C
normoxia controls. Hypoxia is deliberately excluded: its duration design does
not identify `z` well enough for the canonical comparison.

```{r data}
data(zebrafish_o2)
zf <- droplevels(subset(
  zebrafish_o2,
  ploidy == "diploid" & oxygen %in% c("normoxia", "hyperoxia")
))
table(zf$oxygen, zf$temp)
```

```{r standardize}
zf_std <- standardize_data(
  zf, temp = "temp", duration = "duration_min",
  n_total = "n_total", n_surv = "n_surv", duration_unit = "minutes"
)
stopifnot(attr(zf_std, "tdt_meta")$duration_unit == "minutes")
```

## Locked model

The response is beta-binomial. `CTmax`, `z`, and `low` vary by oxygen;
`up` and `k` are shared. Because `t_ref = 60` is expressed in the standardized
minute unit, `CTmax` is the relative-midpoint critical temperature at one hour.

```{r fit}
zf_fit <- fit_4pl(
  zf_std,
  ctmax = ~ 0 + oxygen,
  z = ~ 0 + oxygen,
  low = ~ 0 + oxygen,
  up = ~ 1,
  k = ~ 1,
  family = "beta_binomial",
  t_ref = 60,
  method = "wald",
  quiet = TRUE
)
diagnose_tdt_fit(zf_fit)
```

Always inspect the convergence code, Hessian, gradient, and `check_tls()`
messages before interpretation. This page uses Wald confidence intervals and
names that method explicitly; profile and bootstrap intervals remain available
for sensitivity checks.

```{r quantities}
zf_tls <- tls(zf_fit, by = "oxygen", lethal = FALSE, method = "wald")
zf_table <- zf_tls$summary
names(zf_table)[names(zf_table) == "median"] <- "estimate"
zf_table
```

The reported quantities are `CTmax` and `z`. `Tcrit` is not reported: the
hyperoxia design lacks a benign-temperature control, and the supplement marks
this case `lethal = FALSE` for extraction.

```{r eye, fig.alt="Independently scaled Wald Confidence Eyes for zebrafish CTmax and z under normoxia and hyperoxia, shown as outlined lenses with hollow estimates."}
plot_confidence_eye(zf_fit, parm = c("CTmax", "z"), method = "wald")
```

## Interpretation and boundary

`CTmax` asks whether hyperoxia shifts the one-hour thermal limit. `z` asks
whether oxygen changes the rate at which tolerated duration falls with warming.
Agreement with bayesTLS is a useful cross-check, not proof of correctness:
shared data, filter, or model errors can make both packages agree. Triploid and
hypoxia analyses are not silently substituted here.
