---
title: "Comparing freqTLS to bayesTLS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing freqTLS to bayesTLS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
```

`freqTLS` is the maximum-likelihood / profile-likelihood complement to
[`bayesTLS`](https://github.com/daniel1noble/bayesTLS). This vignette lays out the
relationship, the credit, and a reproducible three-way comparison on the shared
benchmark datasets.

**This vignette builds without Stan.** The live `bayesTLS` (Stan) calls are shown as
display-only R code for reproducibility; the Bayesian and classical two-stage
numbers are read from the version-stamped maintainer-built cache shipped in
`inst/extdata/bayesTLS_benchmark_cache.rds`. The `freqTLS` fits run live. A source
checkout and installed package therefore use the same cached comparator results
without requiring Stan; the recipe below explains how a maintainer rebuilds the
cache from its pinned `bayesTLS` source commit.

```{r setup}
library(freqTLS)
```

## Credit and origins

The thermal-load-sensitivity (TLS) modelling framework implemented here was introduced
by **Daniel W. A. Noble, Pieter A. Arnold, and Patrice Pottier** in the `bayesTLS`
package (manuscript in preparation). The 4PL thermal death-time model and the mapping
from the midpoint slope to `z` and `CTmax` are theirs. `freqTLS` is an independent
likelihood implementation of that framework; please cite `bayesTLS` when you use it.
The bundled `shrimp_lethal` and `zebrafish_lethal` datasets are vendored
from `bayesTLS` under CC BY 4.0 (see `?shrimp_lethal`, `?zebrafish_lethal`, and
`citation("freqTLS")`).

`freqTLS` contributes a TMB maximum-likelihood likelihood, the direct `CTmax`/`z`
reparameterisation that makes both quantities profile-able, and profile-likelihood
confidence intervals.

## The three-way design

The benchmark compares three estimators of the same constant-shape 4PL with a
matched time unit and reference time. The two model fits use the **relative**
midpoint threshold. The classical two-stage estimator uses absolute LT50; it is
shown beside them only for lethal datasets whose fitted asymptotes are near zero
and one, where the two thresholds are close. This is the fairness boundary used
throughout this article:

| Estimator | Path | Uncertainty |
| --- | --- | --- |
| Classical two-stage | `bayesTLS::ts_stage1 -> ts_stage2 -> ts_ci` | delta-method CI |
| Bayesian | `bayesTLS::fit_4pl(temp_effects = "mid") -> extract_tdt(target_surv = "relative")` | posterior credible interval |
| `freqTLS` | `fit_4pl() -> tls(method = "profile")` | profile-likelihood confidence interval |

Under this matched configuration the Bayesian and profile-likelihood fits target the
**same likelihood and the same fitted curve** (see `vignette("model-math")`); they differ
in how they summarise uncertainty — a posterior (with priors) versus a prior-free
likelihood interval.

## The reproducible recipe (not run here)

This is the exact `bayesTLS` recipe used to build the cache on a machine with Stan
installed. It is a plain fenced code block, not an executable vignette chunk.

```r
library(bayesTLS)
data(shrimp_lethal, package = "freqTLS")

# 1. Standardise: name the temperature / duration / count columns and the time
#    unit. fit_4pl() and the two-stage path both consume the standardised frame.
std <- bayesTLS::standardize_data(
  data          = shrimp_lethal,
  temp          = "Temperature_assay",
  duration      = "Duration_exposure_hours",
  n_total       = "N_individuals_after_trial",
  mortality     = "Mortality_after_trial",
  duration_unit = "hours"
)

# 2. Bayesian fit, matched: constant shape via temp_effects = "mid", beta-binomial.
bfit <- bayesTLS::fit_4pl(
  data         = std,
  temp_effects = "mid",
  family       = brms::brmsfamily("beta_binomial", link = "identity"),
  chains       = 4, iter = 4000, seed = 123, backend = "cmdstanr"
)

# 3. Relative-threshold CTmax + z at tref = 1 hour. NOTE: t_ref / time_multiplier
#    live on extract_tdt() (and ts_stage2 / ts_ci), NOT on fit_4pl().
btdt  <- bayesTLS::extract_tdt(
  bfit, target_surv = "relative",
  t_ref = 1, time_multiplier = 1, output_time_unit = "hours"
)
ctmax <- bayesTLS::get_ctmax_summary(btdt)  # temp_lower / temp_median / temp_upper
zsumm <- bayesTLS::get_z_summary(btdt)      # z_median / z_lower / z_upper

# 4. Classical two-stage, same standardised data and reference time.
s1  <- bayesTLS::ts_stage1(std, family = "betabinomial")
s2  <- bayesTLS::ts_stage2(s1, t_ref = 1, time_multiplier = 1)
tci <- bayesTLS::ts_ci(s2, method = "delta", level = 0.95,
                       t_ref = 1, time_multiplier = 1)  # $CTmax_1hr and $z blocks
```

To populate the cache, a maintainer works from a clone of the source repository and
runs the repository-only script `data-raw/build_benchmark_cache.R` on a Stan machine,
with `bayesTLS` installed from a pinned checkout and `BAYESTLS_GIT_SHA` set to its
verified 40-character commit when the installed package lacks `RemoteSha` metadata.
The builder refuses to write a cache if it cannot record that exact source commit.
The script is not installed with the package; it writes
`inst/extdata/bayesTLS_benchmark_cache.rds` (the Bayesian and
two-stage summaries plus a `meta` provenance block: `bayesTLS_version`, `git_sha`, `source_url`,
`cmdstan_version`, `date_built`, `seed`, the configuration, and the data-reconstruction note).

## The freqTLS side (live)

The `freqTLS` fits run live and need nothing beyond this package. We use the
`shrimp_lethal` data (ungrouped) and `zebrafish_lethal` (grouped by life stage). The
shrimp survival counts are reconstructed from the source CSV proportions (see
`?shrimp_lethal`); because all three estimators share the vendored data, they inherit
the same reconstructed counts.

```{r profile-shrimp}
data(shrimp_lethal)
shrimp_std <- standardize_data(
  shrimp_lethal,
  temp = "Temperature_assay", duration = "Duration_exposure_hours",
  n_total = "N_individuals_after_trial", mortality = "Mortality_after_trial",
  duration_unit = "hours"
)
shrimp_fit <- fit_4pl(shrimp_std, t_ref = 1, family = "beta_binomial", quiet = TRUE)
tls(shrimp_fit, method = "profile")$summary
```

`fit_4pl()` returns a `freq_tls` **workflow object** (the twin of bayesTLS's
`bayes_tls`): it bundles the engine fit, the standardised data, and the formula.
Its S3 methods — `tls()`, `confint()`, `summary()`, `plot_confidence_eye()` —
delegate to the engine fit, so you call them on the workflow object directly. The
underlying engine fit is also available as `shrimp_fit$fit` if you want to reach it.

```{r profile-zebrafish}
data(zebrafish_lethal)
zebra_std <- standardize_data(
  zebrafish_lethal,
  temp = "assay_temp", duration = "duration_h",
  n_total = "n_total", n_surv = "n_surv", duration_unit = "hours"
)
zebra_fit <- suppressWarnings(fit_4pl(zebra_std, by = "life_stage",
                                      t_ref = 1, family = "beta_binomial", quiet = TRUE))
# per-stage CTmax and z with profile intervals
tls(zebra_fit, method = "profile")$summary
```

(The zebrafish fit emits a data-adequacy warning about temperatures with fewer than three
durations; this is the kind of identifiability signal `freqTLS` surfaces explicitly. It
is suppressed here only to keep the vignette output tidy.)

## The three-way comparison, with real numbers

The cache holds the maintainer-built `bayesTLS` (posterior) and classical
two-stage summaries; the `freqTLS` column is computed live as this page
renders. The two model fits use the matched beta-binomial, relative-threshold,
constant-shape configuration at `tref = 1` hour, so they target the same fitted
curve. The classical column uses absolute LT50 and is the approximate comparator
described above.

```{r three-way, echo = FALSE}
cache_path <- system.file("extdata", "bayesTLS_benchmark_cache.rds", package = "freqTLS")
if (!nzchar(cache_path) || !file.exists(cache_path)) {
  stop("The shipped bayesTLS benchmark cache is missing; reinstall freqTLS from a complete source tarball.")
}
cache <- readRDS(cache_path)

fmt <- function(est, lo, hi) {
  ifelse(is.na(est), "--", sprintf("%.2f [%.2f, %.2f]", est, lo, hi))
}

pro <- confint(shrimp_fit, c("CTmax", "z"), method = "profile")  # confint() delegates to the engine fit

pick <- function(df, parm, cols) {
  r <- df[df$dataset == "shrimp" & df$parameter == parm, ]
  stats::setNames(as.numeric(r[cols]), c("est", "lo", "hi"))
}
ts_ct <- pick(cache$two_stage, "CTmax", c("estimate", "lower", "upper"))
ts_z  <- pick(cache$two_stage, "z",     c("estimate", "lower", "upper"))
by_ct <- pick(cache$bayesian,  "CTmax", c("median",   "lower", "upper"))
by_z  <- pick(cache$bayesian,  "z",     c("median",   "lower", "upper"))
pr_ct <- pro[pro$parameter == "CTmax", ]
pr_z  <- pro[pro$parameter == "z", ]

knitr::kable(data.frame(
  Quantity = c("CTmax (°C)", "z (°C / decade)"),
  `Two-stage (delta CI)`    = c(fmt(ts_ct["est"], ts_ct["lo"], ts_ct["hi"]),
                                fmt(ts_z["est"],  ts_z["lo"],  ts_z["hi"])),
  `bayesTLS (95% CrI)`      = c(fmt(by_ct["est"], by_ct["lo"], by_ct["hi"]),
                                fmt(by_z["est"],  by_z["lo"],  by_z["hi"])),
  `freqTLS (profile CI)` = c(fmt(pr_ct$estimate, pr_ct$conf.low, pr_ct$conf.high),
                                fmt(pr_z$estimate,  pr_z$conf.low,  pr_z$conf.high)),
  check.names = FALSE
), caption = "Shrimp CTmax and z shown side by side: the two model fits use the relative midpoint; the classical estimator uses absolute LT50.")
```

```{r benchmark-provenance, echo = FALSE}
provenance <- data.frame(
  Field = c("bayesTLS version", "bayesTLS source commit", "CmdStan version",
            "Cache build date", "Seed", "Model threshold", "Shape design"),
  Value = c(
    cache$meta$bayesTLS_version,
    cache$meta$git_sha,
    cache$meta$cmdstan_version,
    cache$meta$date_built,
    cache$meta$seed,
    cache$meta$config$target_surv,
    cache$meta$config$temp_effects
  )
)
knitr::kable(provenance, caption = "Version-stamped benchmark-cache provenance.")
```

For the shrimp data the three estimators land on essentially the same `CTmax` and
`z`, with comparable interval widths: the profile-likelihood confidence
interval and the Bayesian credible interval nearly coincide. That is the point of
the complementary framing — under the matched configuration the likelihood and
the posterior summarise the *same* fitted curve, one with a prior and MCMC, the
other prior-free and by optimisation.

## How fast, accurate, and calibrated

Accuracy and calibration below are descriptive characteristics of the likelihood
path, measured by simulation (the repository-only maintainer script
`data-raw/performance-study.R`, which is not installed with the package) — they ask whether
freqTLS's own intervals are trustworthy, not whether they beat `bayesTLS`
(which buys priors, full posteriors, and the heat-injury sub-models). **Speed**,
though, is one place a head-to-head is both fair and stark.

```{r perf-load, echo = FALSE}
perf_path <- system.file("extdata", "performance_results.rds", package = "freqTLS")
has_perf  <- nzchar(perf_path) && file.exists(perf_path)
if (has_perf) perf <- readRDS(perf_path)
```

**How fast** — a full fit, and one `CTmax` profile interval, by design size:

```{r perf-speed, echo = FALSE}
if (has_perf) {
  knitr::kable(
    perf$speed[, c("family", "design", "n_obs", "median_fit_ms", "median_profile_ms")],
    caption = "Median wall-clock (ms): one fit, and one CTmax profile CI."
  )
}
```

Fits are milliseconds and a profile interval is well under a second. Putting the
three estimators side by side on the shrimp benchmark makes the speed gap concrete:

```{r speed-head-to-head, echo = FALSE}
data("shrimp_lethal", package = "freqTLS", envir = environment())
fmt_t <- function(s) if (is.na(s)) "--" else if (s < 1) sprintf("%.0f ms", s * 1000) else sprintf("%.1f s", s)

sp_std <- standardize_data(shrimp_lethal, temp = "Temperature_assay",
                           duration = "Duration_exposure_hours",
                           n_total = "N_individuals_after_trial",
                           mortality = "Mortality_after_trial", duration_unit = "hours")
t_fit <- system.time(
  sp_fit <- fit_4pl(sp_std, t_ref = 1, family = "beta_binomial", quiet = TRUE)
)[["elapsed"]]
t_wald <- system.time(invisible(confint(sp_fit, c("CTmax", "z"), method = "wald")))[["elapsed"]]
t_prof <- system.time(invisible(confint(sp_fit, c("CTmax", "z"), method = "profile")))[["elapsed"]]

speed <- data.frame(
  Estimator = c("freqTLS", "freqTLS", "freqTLS", "classical two-stage", "bayesTLS"),
  Task = c("fit (ML)", "fit + Wald CTmax & z", "fit + profile CTmax & z",
           "fit + delta CI", "fit (4 chains x 4000 MCMC)"),
  `Wall-clock` = c(fmt_t(t_fit), fmt_t(t_fit + t_wald), fmt_t(t_fit + t_prof), "--", "--"),
  Source = c("live", "live", "live", "cached", "cached"),
  check.names = FALSE, stringsAsFactors = FALSE
)
tpath <- system.file("extdata", "timing_results.rds", package = "freqTLS")
if (nzchar(tpath) && file.exists(tpath)) {
  tm <- readRDS(tpath)$timing
  speed$`Wall-clock`[speed$Estimator == "classical two-stage"] <- fmt_t(tm$seconds[tm$method == "two_stage"])
  speed$`Wall-clock`[speed$Estimator == "bayesTLS"] <- fmt_t(tm$seconds[tm$method == "bayesTLS"])
}
knitr::kable(speed, caption = "Wall-clock on the shrimp benchmark. freqTLS is timed live as this page renders; the two-stage and bayesTLS times are cached from the repository-only maintainer script data-raw/timing-study.R (not installed with the package). The bayesTLS time is post-compile sampling plus overhead, and a first fit also pays a one-time Stan compilation.")
```

freqTLS's **profile** path runs in about a second — comparable to the classical
two-stage and roughly an order of magnitude faster than the Bayesian MCMC fit —
while its **Wald** path is near-instant. That speed is the likelihood path's
concrete advantage, and it is why freqTLS keeps profile the default while
offering Wald as a fast opt-in for well-identified fits. The whole three-way table
above is itself computed live as this page renders, with no Stan.

**How accurate** — bias and RMSE for `CTmax` and `z`:

```{r perf-acc, echo = FALSE}
if (has_perf) {
  knitr::kable(
    perf$accuracy[, c("family", "truth_setting", "parameter", "bias", "rmse", "n_converged")],
    caption = sprintf("Near-unbiased recovery (nsim = %d).", perf$meta$nsim_accuracy)
  )
}
```

**How calibrated** — empirical coverage of the 95% intervals:

```{r perf-cov, echo = FALSE}
if (has_perf) {
  knitr::kable(
    perf$coverage[, c("family", "method", "parameter", "coverage", "median_width", "nominal")],
    caption = sprintf("Profile and Wald 95%% interval coverage (nsim = %d).",
                      perf$meta$nsim_coverage)
  )
}
```

Profile coverage tracks the nominal 95% closely; where it dips (small samples,
beta-binomial `z`), `freqTLS` reports it honestly rather than hiding it.

### Why the beta-binomial profile can dip — and what to use instead

The dip has a specific, diagnosable cause: the dispersion parameter `phi`. When
overdispersion is mild — `phi` large, the data approaching the binomial limit —
`phi` becomes **weakly identified**, and its estimate runs away (its relative
standard error blows up). The profile interval for `CTmax` / `z` profiles that
runaway `phi` out at each grid point, snaps to the binomial limit, and goes too
narrow. The Wald interval propagates the flat-`phi` uncertainty through the joint
Hessian and stays calibrated; the parametric bootstrap is a middle ground. (This
is *not* a clamping artefact — the likelihood's numerical floors never activate
in this regime; see the repository-only maintainer script
`data-raw/beta-binomial-phi-study.R`, which is not installed with the package.)

```{r bb-phi, echo = FALSE}
bbp_path <- system.file("extdata", "beta_binomial_phi_results.rds", package = "freqTLS")
if (nzchar(bbp_path) && file.exists(bbp_path)) {
  bbp <- readRDS(bbp_path)
  z <- bbp$coverage[bbp$coverage$parameter == "z", ]
  phis <- sort(unique(z$phi_true))
  lab <- c("5" = "strong", "50" = "mild", "200" = "very mild")[as.character(phis)]
  pick <- function(m) z$coverage[match(paste(phis, m), paste(z$phi_true, z$method))]
  tab <- data.frame(
    `phi (overdispersion)` = sprintf("%g (%s)", phis, ifelse(is.na(lab), "", lab)),
    profile = pick("profile"), Wald = pick("wald"), bootstrap = pick("bootstrap"),
    check.names = FALSE)
  knitr::kable(tab, digits = 3, caption = sprintf(
    paste("Empirical 95%% coverage of z by method as overdispersion weakens (phi",
          "up = nearer binomial, phi weakly identified; %d sims/cell). The profile",
          "collapses; Wald holds; bootstrap is intermediate."),
    bbp$meta$nsim))
}
```

Accordingly, `fit_tls()` emits an advisory when `phi`'s relative SE is large, and
`confint()` (with the `fallback = TRUE` default) **routes those coordinates to
Wald automatically** — so the default path stays calibrated without you having to
intervene. More broadly, Wald (built on the internal link scale and
back-transformed) is fast and, in these simulations, as well-calibrated as the
profile for fixed effects — a perfectly good choice for routine work. The profile
remains the default because it respects asymmetry without a normal approximation
and now degrades to Wald or the bootstrap exactly where it would be unreliable.

You can see when the automatic routing fires: the `method` column of `confint()`
reads `wald` for the rerouted coordinate even if you asked for `profile`, and an
informational message is emitted. To get an asymmetry-respecting interval for that
coordinate specifically, request the bootstrap — for example
`confint(fit, "phi", method = "bootstrap")`.

**When to use each method:**

| Method | Use when |
| --- | --- |
| `profile` (default) | the headline choice — respects asymmetry, no normal approximation, and degrades honestly when a coordinate is weakly identified. |
| `wald` | fast routine work; as well-calibrated as the profile for fixed effects in these simulations; symmetric on the link scale. |
| `bootstrap` | a prior-free fallback that can provide an asymmetry-respecting interval for `up`, a weak `phi`, or another weakly identified coordinate; failed or degenerate replicates can still leave an open or unavailable interval. |

## Bootstrap fallback when a profile stays open

When a profile does not close — for example under a weakly identified design or
a boundary asymptote — `confint()` can fall back to a prior-free **parametric
bootstrap**. The fallback often yields an asymmetry-respecting interval, but it
does not manufacture certainty: too few valid or non-degenerate bootstrap fits
can still produce an unavailable bound. Set `fallback = FALSE` to retain the
open profile explicitly.

```{r bootstrap-demo}
# A deliberately sparse design where the CTmax profile does not close.
sparse <- simulate_tls(family = "binomial", temps = c(35, 36), times = c(1, 2),
                       reps = 2, n = 10, CTmax = 36, z = 4, seed = 9)
sfit_sparse <- suppressWarnings(fit_tls(sparse, y = survived, n = total,
                                        time = duration, temp = temp,
                                        family = "binomial", tref = 1))
# Strict profile: NA on an open side. Default: a parametric bootstrap interval.
strict <- suppressWarnings(
  confint(sfit_sparse, "CTmax", method = "profile", fallback = FALSE))
boot <- suppressWarnings(
  confint(sfit_sparse, "CTmax", method = "profile", nboot = 1000, boot_seed = 1))
data.frame(
  setting   = c("strict profile (fallback = FALSE)", "default (bootstrap fallback)"),
  conf.low  = c(strict$conf.low,  boot$conf.low),
  conf.high = c(strict$conf.high, boot$conf.high),
  method    = c(strict$method,    boot$method)
)
```

When enough bootstrap refits remain stable and non-degenerate, `freqTLS` can
provide a prior-free interval after an open profile. Sparse or boundary designs
can still leave a bound unavailable; the status column reports that outcome.

## The teaching device: posterior density versus Confidence Eye

The clearest way to *see* the Bayesian-versus-likelihood distinction is to draw, for the
same `CTmax` (or `z`):

* the **`bayesTLS` posterior density** — a probability distribution over the parameter,
  shaped by the prior and the data; and
* the **`freqTLS` Confidence Eye** — a confidence lens with a hollow point estimate,
  carrying no prior and making no probability statement about the parameter.

```{r eye, fig.alt = "Confidence Eye for the shrimp CTmax and z: pale confidence lenses with hollow point estimates, the freqTLS uncertainty display."}
plot_confidence_eye(shrimp_fit, parm = c("CTmax", "z"), method = "profile")
```

The Confidence Eye above is the `freqTLS` half of that contrast. When the cache (and
`bayesTLS`) are available, the posterior density for the same quantity can be drawn beside
it; the side-by-side makes explicit that one is a posterior and the other is a likelihood
confidence interval. `freqTLS` deliberately never renders a posterior-style density
for its own intervals, and its prose uses "confidence" language, never
"posterior" or "credible".

## Beyond the matched shape: stage-specific curves

The three-way comparison holds the shape constant (`low`, `up`, `k` shared across
temperatures and groups) so all three estimators target the *same* curve — that
is what makes the benchmark fair. `freqTLS` can also relax that restriction: the
shape parameters `low`, `up`, and `log_k` may vary by a grouping factor. Re-fitting `zebrafish_lethal` with stage-specific shapes and comparing by
AIC asks whether the life stages differ in more than thermal *location*
(`CTmax`, `z`):

```{r v02-stage-shape}
stage_shape <- suppressWarnings(fit_4pl(
  zebra_std, by = "life_stage",
  low = ~ life_stage, up = ~ life_stage, k = ~ life_stage,
  t_ref = 1, family = "beta_binomial", quiet = TRUE
))
# zebra_fit (above) is the shared-shape fit; stage_shape lets low / up / k vary.
c(shared_shape_AIC = round(AIC(zebra_fit), 1),
  stage_shape_AIC  = round(AIC(stage_shape), 1))
```

The stage-specific model has the substantially lower AIC, so the data support
per-stage shapes. The difference is concentrated in the upper asymptote `up` (the
maximum survival at benign exposures), not the steepness `k`:

```{r v02-up}
stage_est <- stage_shape$fit$estimates
stage_est[grepl("^up:", stage_est$parameter), c("parameter", "estimate", "std.error")]
```

Young embryos have a markedly lower survival ceiling (`up` near 0.7) than older
embryos and larvae (near 0.9) — a real biological difference that the matched
constant-shape configuration used for the `freqTLS` and `bayesTLS` model-based
benchmark cannot express. The classical two-stage workflow is a separate
absolute-LT50 approximation and likewise does not model a stage-specific 4PL
shape.

`freqTLS` also reads absolute critical temperatures and predicts heat injury
off the same fitted curve. For the (ungrouped) shrimp fit, the absolute critical
temperature for 50% survival at one hour, and the survival predicted under a
four-hour exposure at 32 °C:

```{r v02-derive}
c(
  CTmax_50pct_1h = round(derive_ctmax(shrimp_fit$fit, surv = 0.5, duration = 1), 2),
  surv_32C_4h    = round(tail(predict_heat_injury(shrimp_fit$fit,
                     data.frame(time = seq(0, 4, by = 0.1), temp = 32))$survival, 1), 3)
)
```

(`derive_tcrit()` similarly returns a rate-multiplier `T_crit`, with an explicit
lethal-endpoint caveat.) These are deterministic transforms of the fitted
`CTmax` / `z`, not new fits.

## When to prefer which

* Use `freqTLS` when you want fast, prior-free, asymmetry-respecting intervals and an
  explicit identifiability check. When the profile does not close it falls back to a
  parametric bootstrap; unstable or degenerate refits can still leave the interval
  unavailable.
* Use `bayesTLS` when you want a full Bayesian workflow, prior information, or the
  heat-injury and repair sub-models.

The two are complementary lenses on the same model. `freqTLS` flags weak
identifiability explicitly and never claims the profile is universally superior; see
`vignette("profile-likelihood")` for the non-closing behaviour and the bootstrap
fallback.
