---
title: "Getting started with freqTLS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with freqTLS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
```

`freqTLS` fits single-stage four-parameter logistic (4PL) thermal-load-sensitivity
(TLS) / thermal death-time models by maximum likelihood, parameterised **directly in
`CTmax` and `z`** (thermal sensitivity — the temperature rise that cuts the tolerated
exposure roughly tenfold), and returns prior-free **frequentist** confidence
intervals — Wald, profile-likelihood (asymmetry-respecting, the default), and bootstrap. It is the likelihood complement to the
Bayesian [`bayesTLS`](https://github.com/daniel1noble/bayesTLS) package; the modelling
framework is theirs (see `vignette("comparing-to-bayesTLS")` for the relationship and
credit).

This vignette walks through the core loop: simulate → fit → confidence intervals →
plot. It runs end to end with a small simulation and needs no Stan, no MCMC, and no
internet.

```{r setup}
library(freqTLS)
```

## Simulate a dataset

`simulate_tls()` draws survival counts from the locked data-generating process: a
factorial grid of assay temperatures and exposure durations, with the 4PL evaluated at
the supplied true `CTmax`, `z`, and shape parameters. We use the overdispersed
beta-binomial family here; `phi` is its precision parameter — larger `phi` means
*less* overdispersion, approaching the ordinary binomial as `phi` grows.

```{r simulate}
set.seed(1)
dat <- simulate_tls(
  temps  = seq(30, 42, by = 2),
  times  = c(0.5, 1, 2, 4, 8),
  reps   = 3,
  n      = 20,
  CTmax  = 36,
  z      = 4,
  phi    = 50,
  family = "beta_binomial",
  seed   = 1
)
head(dat)
```

Each row is one temperature-by-duration cell: `survived` out of `total` individuals
survived the exposure. The true parameters used to generate the data are kept as an
attribute, which is handy for checking recovery:

```{r truth}
attr(dat, "truth")[c("CTmax", "z", "phi")]
```

## Standardise and fit the model

`freqTLS` mirrors the `bayesTLS` workflow. First `standardize_data()` names the
temperature, duration, and survival-count columns and records the data contract;
then `fit_4pl()` fits the 4PL by maximum likelihood and returns a `freq_tls`
workflow object. `t_ref` is the reference time at which `CTmax` is defined: the
shared-name facade `fit_4pl()` uses the `bayesTLS` spelling `t_ref` (default 60), while
the lower-level engine `fit_tls()` (below) uses `tref` (default 1) — the same
quantity, in the data's duration unit. The simulated durations here are already in
reference units, so we pass `1`.

```{r standardize}
std <- standardize_data(
  dat,
  temp     = "temp",
  duration = "duration",
  n_total  = "total",
  n_surv   = "survived"
)
```

```{r fit}
fit <- fit_4pl(std, family = "beta_binomial", t_ref = 1)
fit
```

### Formula interface

For the supported moderators, `fit_4pl()` takes a **similar direct interface** to `bayesTLS` —
`ctmax =`, `z =`, or the `by =` shorthand (e.g. `by = "population"` for a grouped
fit). Under the hood the engine also exposes a `brms`/`drmTMB`-style grammar via
`tls_bf()` — the left-hand side names the survival counts (`successes |
trials(total)`), the right-hand side tags the two axes with the `time()` and
`temp()` markers — which feeds the same likelihood, so the two fits are
numerically identical:

```{r fit-formula}
fit_f <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp)),
  data   = dat,
  family = "beta_binomial",
  tref   = 1
)
all.equal(coef(fit_f), coef(fit))
```

The grammar scales well beyond a single ungrouped fit: each sub-parameter can take
a grouping factor, a continuous covariate, or an `lme4`-style random intercept.
See **Going further** below.

The print method reports the headline quantities (`CTmax`, `z`), the shape parameters
(`low`, `up`, `k`), the overdispersion `phi`, and the convergence status. `summary()`
adds standard errors and the log-likelihood:

```{r summary}
summary(fit)
```

A tidy, broom-style table of all natural-scale parameters is available with
`tidy_parameters()`, and the headline quantities have dedicated extractors:

```{r tidy}
tidy_parameters(fit)
get_ctmax(fit)
get_z(fit)
```

## Profile-likelihood confidence intervals

The reason for the direct `CTmax`/`z` parameterisation is that both headline quantities
are model *coordinates*, so they can be profiled directly. `confint()` with
`method = "profile"` (the default) inverts the likelihood-ratio test: the interval is
the set of values whose deviance from the maximum stays below the profile-$t$ cutoff (a
squared Student-$t$ quantile on the residual degrees of freedom, not $\chi^2_1$; see
`vignette("frequentist-and-bayesian")`), found by root-finding on each side of the MLE.

```{r confint}
confint(fit, parm = "CTmax", method = "profile")
confint(fit, parm = "z", method = "profile")
```

These are **confidence** intervals, not posteriors: they carry no prior, and they
need not be symmetric about the estimate. The `conf.status` column reports `"ok"` when
the profile closes on both sides; a weakly identified parameter is flagged rather than
given a fabricated bound (see `vignette("profile-likelihood")`).

For comparison, the first-order Wald interval is available with `method = "wald"`:

```{r wald}
confint(fit, parm = c("CTmax", "z"), method = "wald")
```

## Plot the fit

`plot_survival_curves()` draws the fitted survival probability against exposure
duration, one curve per temperature, with the observed proportions overlaid:

```{r survival-curves, fig.alt = "Fitted 4PL survival curves: survival probability declining with exposure duration, one curve per assay temperature, with observed proportions as points."}
plot_survival_curves(fit)
```

The default uncertainty display is the **Confidence Eye**: a pale confidence lens
with a hollow point estimate. It is freqTLS's visual identity and deliberately avoids
posterior-density iconography, because these are likelihood intervals. Read the lens
**width** as the confidence interval and the **hollow circle** as the
maximum-likelihood estimate; two eyes that clearly fail to overlap are
distinguishable at the confidence level, and a hollow point with *no* lens flags a
profile that did not close (a weakly identified parameter).

```{r confidence-eye, fig.alt = "Confidence Eye plot: a pale lens spanning the profile-likelihood interval with a hollow point estimate, for CTmax and z."}
plot_confidence_eye(fit, parm = c("CTmax", "z"), method = "profile")
```

## Going further: groups, shape predictors, and random effects

The same `tls_bf()` grammar scales from the single fit above to grouped,
covariate, and hierarchical models.

**Other families.** Besides `"beta_binomial"`, `fit_tls()` accepts `"binomial"`
(no overdispersion) and `"beta"` (a continuous proportion in (0, 1) with no trials
column; the runnable example below uses simulated beta data).

**Grouped `CTmax` and `z`.** Put a grouping factor on a sub-parameter to estimate a
separate, directly profile-able `CTmax` and `z` per group, with one shared curve
shape (the `bayesTLS` constant-shape configuration). In the formula, `z` enters on
its internal log scale as `log_z` (and steepness as `log_k`), so the grouped term is
written `log_z ~ group`. Intervals are still requested by the natural-scale name
(`z:cool`, not `log_z:cool`) — `confint()` converts internally:

```{r grouped}
dat_grp <- simulate_tls(
  temps = seq(30, 42, by = 2), times = c(0.5, 1, 2, 4, 8), reps = 3, n = 20,
  group = c("cool", "warm"), CTmax = c(34, 38), z = c(3, 5),
  phi = 50, family = "beta_binomial", seed = 2
)
fit_grp <- fit_tls(
  tls_bf(
    survived | trials(total) ~ time(duration) + temp(temp),
    CTmax ~ group,
    log_z ~ group
  ),
  data = dat_grp, family = "beta_binomial", tref = 1
)
confint(fit_grp, c("CTmax:cool", "CTmax:warm", "z:cool", "z:warm"),
        method = "profile")
```

**Shape predictors.** The shape parameters are no longer intercept-only: `low`,
`up`, and `log_k` each take their own grouping factor or continuous covariate
(for example `low ~ group` or `log_k ~ body_size`). See `?tls_bf`.

**A random intercept.** Add an `lme4`-style `(1 | group)` term to let a
sub-parameter vary across groups drawn from a population — here `CTmax` varies by
`colony`:

```{r random-effect}
dat_re <- simulate_tls(
  temps = seq(30, 42, by = 2), times = c(0.5, 1, 2, 4, 8), reps = 2, n = 20,
  CTmax = 36, z = 4, phi = 50, family = "beta_binomial",
  re_sd = 1.2, n_re_groups = 10, seed = 3
)
fit_re <- fit_tls(
  tls_bf(
    survived | trials(total) ~ time(duration) + temp(temp),
    CTmax ~ (1 | colony)
  ),
  data = dat_re, family = "beta_binomial", tref = 1
)
fit_re
head(ranef(fit_re))
```

The intercept is integrated out by a Laplace approximation; `sigma_CTmax` is its
standard deviation and `ranef()` returns the per-colony BLUPs. The population
`CTmax` interval is profiled under the random effect (`confint(fit_re, "CTmax")`,
the default), which propagates the variance-component uncertainty that a Wald
interval would understate. Single random intercepts are supported on `CTmax`,
`log_z`, `low`, and `log_k`; see `vignette("random-effects")` for the full
hierarchical workflow and the few-groups caveats (the variance components are
biased low when fewer than about eight groups are sampled).

## The core workflow at a glance

`freqTLS` deliberately mirrors the `bayesTLS` workflow, so the same
**standardise → fit → quantities → plot** path runs by maximum likelihood.
The map below shows the main workflow and how each box corresponds to a `bayesTLS`
function. The
**"freqTLS extras"** box (gold, upper right) is the frequentist-only addition —
Wald / profile / bootstrap intervals and the Confidence Eye — and the dashed
**"trace & repair helpers — planned"** box is the one piece not yet ported.

```{r function-map, echo = FALSE, results = "asis"}
# Inline the SVG rather than include_graphics() so the figure does not depend on
# pkgdown copying a co-located static asset. The Pandoc raw-HTML fence is
# essential: without it, wildcard labels such as get_*_summary() are parsed as
# Markdown emphasis, which breaks the SVG namespace. Accessibility is provided
# by the SVG's <title>/<desc>.
svg_candidates <- c(
  "freqTLS_function_map.svg",
  system.file("doc", "freqTLS_function_map.svg", package = "freqTLS")
)
svg_path <- svg_candidates[nzchar(svg_candidates) & file.exists(svg_candidates)][1]
if (is.na(svg_path)) {
  stop("The installed freqTLS function-map SVG is missing.")
}
svg_lines <- readLines(svg_path, warn = FALSE)
cat(
  "```{=html}\n",
  paste(svg_lines, collapse = "\n"),
  "\n```\n",
  sep = ""
)
```

## Where to next

* `vignette("model-math")` — the 4PL, the direct `CTmax`/`z` parameterisation, and the
  exact bridge to `bayesTLS`.
* `vignette("profile-likelihood")` — what the profile is doing, why intervals can be
  asymmetric, profile versus Wald, and the honest non-closing fallback.
* `vignette("random-effects")` — hierarchical thermal-tolerance models: random
  intercepts on `CTmax`, `log_z`, `low`, and `log_k`, with their diagnostics.
* `vignette("comparing-to-bayesTLS")` — the three-way comparison (classical two-stage,
  Bayesian, profile likelihood) and the credit for the framework.

Before interpreting any fitted model, run `check_tls(fit)`. The recovery guide in
`?check_tls` maps each warning to the next design or analysis action; the profile
article shows both the default bootstrap recovery attempt and the strict
`fallback = FALSE` open-profile diagnostic.
