---
title: "Experimental extension: deterministic heat-injury prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental extension: deterministic heat-injury prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

This page is a freqTLS-only experimental extension built from synthetic data.
It is not a replacement empirical case and does not estimate a repair process.

```{r fit}
library(freqTLS)
sim <- simulate_tls(family = "beta_binomial", CTmax = 36, z = 4,
                    phi = 50, seed = 42)
# In this synthetic example, duration and trace time are both hours. No unit
# conversion occurs inside fit_tls() or predict_heat_injury().
fit <- fit_tls(sim, y = survived, n = total, time = duration, temp = temp,
               family = "beta_binomial", tref = 1, quiet = TRUE)
diagnose_tdt_fit(fit)
```

```{r trace}
trace <- data.frame(
  time = seq(0, 96, by = 0.25)
)
trace$temp <- 27 + 6 * sin(2 * pi * (trace$time - 6) / 24)
```

Here `tref = 1` is one hour, and `trace$time` is also in hours. The trace time,
fitted duration, and reference time must use the same unit because heat-injury
damage is accumulated per time unit.

```{r prediction}
injury <- predict_heat_injury(fit, trace)
head(injury)
```

```{r plot, fig.alt="Deterministic survival trajectory under a synthetic fluctuating temperature trace."}
plot_heat_injury(fit, trace, nboot = 50, seed = 42)
```

The trajectory is a deterministic transformation of the fitted 4PL and the
user-supplied trace. `heat_injury_envelope()` can add parametric-bootstrap
uncertainty from refitted curves. A repair rate supplied by the user remains an
illustrative scenario: freqTLS has not fitted that rate from damage/recovery
data. Use bayesTLS for its broader Bayesian heat-injury and repair workflow; its
posterior intervals are distinct from freqTLS confidence intervals.
