---
title: "Experimental extension: deterministic heat-injury prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental extension: deterministic heat-injury prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
```

For diploid zebrafish larvae under normoxia, how would a fitted static-assay
survival curve translate into predicted survival during a hypothetical
three-hour staged heat exposure?
This page uses the package's CC BY 4.0 oxygen-gradient zebrafish assay data and
a clearly hypothetical exposure scenario. It is a freqTLS-only experimental
extension: it does not estimate injury kinetics or a repair process.

```{r fit}
library(freqTLS)
data("zebrafish_o2")
zebrafish_normoxia <- subset(
  zebrafish_o2,
  ploidy == "diploid" & oxygen == "normoxia"
)
# The assay duration, reference time, and scenario trace below are all minutes.
# `tref = 60` therefore defines CTmax at one hour.
fit <- fit_tls(
  zebrafish_normoxia,
  y = n_surv, n = n_total, time = duration_min, temp = temp,
  family = "beta_binomial", tref = 60, quiet = TRUE
)
diagnose_tdt_fit(fit)
```

```{r trace}
trace <- data.frame(
  time = seq(0, 180, by = 5)
)
trace$temp <- ifelse(trace$time <= 60, 38,
                     ifelse(trace$time <= 120, 39, 40))
```

The staged 38–40 °C heatwave is a user-defined scenario, not an environmental
trace. Those temperatures and the 20.9–240 minute duration range occur in the
static assay, but their fluctuating sequence was not observed. The trace time,
fitted duration, and reference time must use the same unit because heat-injury
damage is accumulated per time unit.

```{r prediction}
injury <- predict_heat_injury(fit, trace)
head(injury)
```

```{r plot, fig.alt="Predicted survival for a hypothetical staged heatwave applied to the fitted diploid-normoxia zebrafish survival curve."}
plot_heat_injury(fit, trace, nboot = 50, seed = 42, xlab = "Time (minutes)")
```

`nboot = 50` keeps this vignette quick to render; use substantially more
bootstrap replicates before reporting a confidence band.

The default one-dose survival target is the fitted relative midpoint, defined
from `low` and `up`; it is not an absolute 50% survival threshold. To inspect a
separate absolute-threshold scenario, use `target_surv = 0.5` only when 0.5 is
strictly between the fitted asymptotes. That sensitivity analysis is not CTmax.

The trajectory is a deterministic transformation of the fitted 4PL and the
user-supplied trace. `heat_injury_envelope()` adds parametric-bootstrap
confidence bands from refitted curves. Variable-temperature dose additivity is
an extrapolation from the static assays, and a repair rate supplied by the user
remains an illustrative scenario: freqTLS has not fitted injury or repair
parameters from damage/recovery data. Use bayesTLS when fitted heat-injury or
repair dynamics, or Bayesian scenario propagation, is the scientific aim; its
posterior intervals are distinct from freqTLS confidence intervals.
