---
title: "Profile-likelihood intervals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile-likelihood intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
```

`freqTLS` reports **profile-likelihood confidence intervals** as its default. This
vignette explains what the profile is, why its intervals can be asymmetric, how it
differs from the Wald interval, and — importantly — how `freqTLS` behaves honestly
when a profile does not close. These are likelihood intervals, not posteriors; the
language throughout is "confidence", never "posterior" or "credible".
The authoritative algorithm reference is `docs/design/04-profile-likelihood.md`.

```{r setup}
library(freqTLS)
```

## What the profile does

For a scalar target $\psi$ (say `CTmax`):

1. Fit the maximum-likelihood estimate, obtaining $(\hat\theta, \hat\ell)$.
2. Fix the internal coordinate that maps to $\psi$ at a candidate value and re-optimise
   all the other coordinates, giving the profile log-likelihood $\ell_p(\psi)$.
3. Form the deviance $D(\psi) = 2\,(\hat\ell - \ell_p(\psi))$.
4. The confidence interval is the set $\{\psi : D(\psi) \le t^2_{1-\alpha/2,\,\nu}\}$,
   found by root-finding on each side of the MLE.

At the MLE the deviance is (numerically) zero. The cutoff is the **squared
Student-$t$ quantile** $t^2_{1-\alpha/2,\,\nu}$ on $\nu = n - p$ residual degrees of
freedom (data rows minus free fixed-effect coordinates), not the $\chi^2_1$ quantile:
this is the Bates–Watts profile-$t$ calibration, which is less optimistic at small $n$
and converges to $\chi^2_1$ as $\nu \to \infty$ (see `vignette("frequentist-and-bayesian")`).
`profile()` returns the whole deviance curve plus the interval:

```{r profile-object}
set.seed(1)
dat <- simulate_tls(family = "beta_binomial", CTmax = 36, z = 4, phi = 50, seed = 1)
fit <- fit_tls(dat, y = survived, n = total, time = duration, temp = temp,
               family = "beta_binomial", tref = 1)

pc <- profile(fit, "CTmax")
c(estimate = pc$estimate, conf.low = pc$conf.low, conf.high = pc$conf.high,
  cutoff = pc$cutoff, min_deviance = min(pc$deviance))
```

The deviance minimum sits essentially at zero, at the estimate. Plotting the profile
shows the deviance curve, the profile-$t$ cutoff line, and the interval where the curve
dips below it:

```{r profile-plot, fig.alt = "Profile-likelihood deviance curve for CTmax: a U-shaped curve with a horizontal profile-t cutoff line; the confidence interval is where the curve lies below the cutoff."}
plot(pc)
```

## Asymmetry and equivariance

A profile interval need not be symmetric about the estimate, and `freqTLS` preserves
the asymmetry rather than forcing a symmetric $\pm$ band. For a parameter on a log scale,
such as `z`, the profile is taken on the internal `log_z` coordinate and the endpoints
are exponentiated. This makes the interval **equivariant**: the `z` interval is exactly
$\exp()$ of the internal `log_z` interval.

```{r equivariance}
ci_z     <- confint(fit, "z",     method = "profile")
ci_log_z <- confint(fit, "log_z", method = "profile")

# z endpoints equal exp() of the log_z endpoints
rbind(
  z          = c(ci_z$conf.low, ci_z$conf.high),
  exp_log_z  = exp(c(ci_log_z$conf.low, ci_log_z$conf.high))
)

# the interval is asymmetric about the estimate (in general)
with(ci_z, c(lower_gap = estimate - conf.low, upper_gap = conf.high - estimate))
```

## Profile versus Wald

The Wald interval is $\hat\psi \pm z_{\alpha/2}\,\mathrm{se}$ computed on the internal
(link) scale and back-transformed. It is fast and first-order, but symmetric on the link
scale and blind to the curvature of the likelihood. The profile interval inverts the
likelihood-ratio test directly. On well-identified data the two agree closely; they
diverge when the likelihood is skewed or the estimate approaches a boundary.

```{r profile-vs-wald}
rbind(
  profile = unlist(confint(fit, "CTmax", method = "profile")[c("conf.low", "conf.high")]),
  wald    = unlist(confint(fit, "CTmax", method = "wald")[c("conf.low", "conf.high")])
)
```

`tidy_parameters()` switches between the two for the whole parameter table:

```{r tidy-methods}
tidy_parameters(fit, method = "profile")[, c("parameter", "estimate", "conf.low", "conf.high", "interval_type")]
```

### The upper asymptote `up`

Under the disjoint-bounds parameterisation `up` has its own coordinate `beta_up`, but
`freqTLS` does not yet profile it (the profile path is wired for `low` but not `up`).
`freqTLS` reports `up` with the delta-method
Wald interval and labels it honestly — `confint(fit, "up", method = "profile")` returns
`interval_type = "wald"` and emits an informational message rather than silently
substituting a different quantity.

## The honest non-closing fallback

The headline value-add over the Bayesian path is that `freqTLS` tells you when the
data do not identify a parameter, instead of letting a prior quietly fill the gap. When a
profile does not rise above the cutoff on one side — because the data are too sparse to
pin the parameter down — `freqTLS`:

* emits a warning that the parameter is **weakly identified** ("consider `bayesTLS` or a
  bootstrap"),
* returns `NA` on the open side (never a fabricated bound), and
* sets a `conf.status` marker (`open_lower`, `open_upper`, or `open_both`).

Here is a deliberately sparse design — few temperatures, almost no mortality contrast —
that does not identify `CTmax`. We catch the warning so the vignette builds, but show the
status and the `NA` endpoint:

```{r non-closing}
set.seed(7)
sparse <- simulate_tls(
  temps  = c(35, 36),     # only two temperatures
  times  = c(1, 2),       # only two durations
  reps   = 2, n = 8,
  CTmax  = 36, z = 4,
  family = "binomial", seed = 7
)
sparse_fit <- suppressWarnings(
  fit_tls(sparse, y = survived, n = total, time = duration, temp = temp,
          family = "binomial", tref = 1)
)

ci <- tryCatch(
  withCallingHandlers(
    confint(sparse_fit, "CTmax", method = "profile"),
    warning = function(w) {
      message("caught warning: ", conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  ),
  error = function(e) e
)
ci[, c("parameter", "conf.low", "conf.high", "estimate", "conf.status")]
```

The interval is open (an `NA` endpoint and an `open_*` status) rather than a confident
but unsupported bound. The Confidence Eye honours this too: a non-closing profile draws a
**hollow point with no lens**, never a fabricated closed eye.

```{r non-closing-eye, fig.alt = "Confidence Eye for a weakly identified fit: hollow point estimates with no confidence lens, signalling that the profile did not close."}
suppressWarnings(plot_confidence_eye(sparse_fit, parm = "CTmax", method = "profile"))
```

## Calibration: how well do the intervals cover?

A confidence interval is only as good as its coverage. `data-raw/coverage-study.R`
simulates 200 datasets at a known `CTmax`/`z` under each family, fits every one by ML,
builds 95% profile intervals, and records the empirical (frequentist) coverage. The
summary is shipped with the package:

```{r coverage}
cov_path <- system.file("extdata", "coverage_results.rds", package = "freqTLS")
if (nzchar(cov_path)) {
  cov <- readRDS(cov_path)
  knitr::kable(
    cov$coverage, digits = 3,
    caption = sprintf("Empirical coverage of 95%% profile CIs (nsim = %d; nominal 0.95).",
                      cov$meta$nsim)
  )
}
```

For the **binomial** model the profile intervals are well calibrated — `CTmax` coverage
sits at the nominal 0.95 and every profile closed. For the **beta-binomial** model the
intervals **under-cover** (`CTmax` ≈ 0.84, `z` ≈ 0.87) at this sample size: estimating the
extra overdispersion parameter makes the likelihood-ratio interval optimistic, and a small
fraction of profiles do not close. The honest reading is to treat beta-binomial profile
intervals as approximate and, when coverage matters, calibrate them with a parametric
bootstrap or use `bayesTLS`. This is exactly the regime the ship stance below warns about.
Re-run `data-raw/coverage-study.R` to reproduce (and to regenerate the per-simulation raw
data).

## Ship stance

The profile gives fast, prior-free, asymmetry-respecting confidence intervals **when
the MLE is interior and the data identify the target**. For boundary asymptotes, very
sparse designs, overdispersion concentrated at zero, or (future) random effects, prefer
`bayesTLS` or a bootstrap — and `freqTLS` warns you when you are in that regime. It
never claims the profile is universally superior to the Bayesian path. See
`vignette("comparing-to-bayesTLS")` for the side-by-side comparison.
