---
title: "Profile-likelihood intervals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Profile-likelihood intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
```

`freqTLS` reports **profile-likelihood confidence intervals** as its default. This
vignette explains what the profile is, why its intervals can be asymmetric, how it
differs from the Wald interval, and — importantly — how `freqTLS` behaves honestly
when a profile does not close. These are likelihood intervals, not posteriors; the
language throughout is "confidence", never "posterior" or "credible". See
`vignette("model-math")` for the 4PL parameterisation that the profile follows.
Before interpreting an interval, run `check_tls(fit)`; its help page maps each
data-adequacy warning to a concrete design or analysis response.

```{r setup}
library(freqTLS)
```

## What the profile does

For a scalar target $\psi$ (say `CTmax`):

1. Fit the maximum-likelihood estimate, obtaining $(\hat\theta, \hat\ell)$.
2. Fix the internal coordinate that maps to $\psi$ at a candidate value and re-optimise
   all the other coordinates, giving the profile log-likelihood $\ell_p(\psi)$.
3. Form the deviance $D(\psi) = 2\,(\hat\ell - \ell_p(\psi))$.
4. The confidence interval is the set $\{\psi : D(\psi) \le t^2_{1-\alpha/2,\,\nu}\}$,
   found by root-finding on each side of the MLE.

At the MLE the deviance is (numerically) zero. The cutoff is the **squared
Student-$t$ quantile** $t^2_{1-\alpha/2,\,\nu}$ on $\nu = n - p$ residual degrees of
freedom (data rows minus free fixed-effect coordinates), not the $\chi^2_1$ quantile:
this is the Bates–Watts profile-$t$ calibration, which is less optimistic at small $n$
and converges to $\chi^2_1$ as $\nu \to \infty$ (see `vignette("frequentist-and-bayesian")`).
`profile()` returns the whole deviance curve plus the interval:

```{r profile-object}
set.seed(1)
dat <- simulate_tls(family = "beta_binomial", CTmax = 36, z = 4, phi = 50, seed = 1)
fit <- fit_tls(dat, y = survived, n = total, time = duration, temp = temp,
               family = "beta_binomial", tref = 1)

pc <- profile(fit, "CTmax")
c(estimate = pc$estimate, conf.low = pc$conf.low, conf.high = pc$conf.high,
  cutoff = pc$cutoff, min_deviance = min(pc$deviance))
```

The deviance minimum sits essentially at zero, at the estimate. Plotting the profile
shows the deviance curve, the profile-$t$ cutoff line, and the interval where the curve
dips below it:

```{r profile-plot, fig.alt = "Profile-likelihood deviance curve for CTmax: a U-shaped curve with a horizontal profile-t cutoff line; the confidence interval is where the curve lies below the cutoff."}
plot(pc)
```

## Asymmetry and equivariance

A profile interval need not be symmetric about the estimate, and `freqTLS` preserves
the asymmetry rather than forcing a symmetric $\pm$ band. For a parameter on a log scale,
such as `z`, the profile is taken on the internal `log_z` coordinate and the endpoints
are exponentiated. This makes the interval **equivariant**: the `z` interval is exactly
$\exp()$ of the internal `log_z` interval.

```{r equivariance}
ci_z     <- confint(fit, "z",     method = "profile")
ci_log_z <- confint(fit, "log_z", method = "profile")

# z endpoints equal exp() of the log_z endpoints
rbind(
  z          = c(ci_z$conf.low, ci_z$conf.high),
  exp_log_z  = exp(c(ci_log_z$conf.low, ci_log_z$conf.high))
)

# the interval is asymmetric about the estimate (in general)
with(ci_z, c(lower_gap = estimate - conf.low, upper_gap = conf.high - estimate))
```

## Profile versus Wald

The Wald interval is $\hat\psi \pm z_{\alpha/2}\,\mathrm{se}$ computed on the internal
(link) scale and back-transformed. It is fast and first-order, but symmetric on the link
scale and blind to the curvature of the likelihood. The profile interval inverts the
likelihood-ratio test directly. On well-identified data the two agree closely; they
diverge when the likelihood is skewed or the estimate approaches a boundary.

```{r profile-vs-wald}
rbind(
  profile = unlist(confint(fit, "CTmax", method = "profile")[c("conf.low", "conf.high")]),
  wald    = unlist(confint(fit, "CTmax", method = "wald")[c("conf.low", "conf.high")])
)
```

`tidy_parameters()` switches between the two for the whole parameter table:

```{r tidy-methods}
tidy_parameters(fit, method = "profile")[, c("parameter", "estimate", "conf.low", "conf.high", "interval_type")]
```

### The upper asymptote `up`

Under the disjoint-bounds parameterisation `up` has its own coordinate `beta_up`, but
`freqTLS` does not yet profile it (the profile path is wired for `low` but not `up`).
`freqTLS` reports `up` with the delta-method
Wald interval and labels it honestly — `confint(fit, "up", method = "profile")` returns
`interval_type = "wald"` and emits an informational message rather than silently
substituting a different quantity. If you need a prior-free, asymmetry-respecting
interval for `up`, request the bootstrap: `confint(fit, "up", method = "bootstrap")`.

## Recovery when a profile does not close

The headline value-add over the Bayesian path is that `freqTLS` tells you when the
data do not identify a parameter, instead of letting a prior quietly fill the gap. When a
profile does not rise above the cutoff on one side — because the data are too sparse to
pin the parameter down — the default `fallback = TRUE` first attempts a parametric
bootstrap. A successful bootstrap returns a prior-free percentile interval and labels
its `method` as `"bootstrap"`; too few stable or non-degenerate refits still leave an
unavailable interval. Set `fallback = FALSE` when you specifically need to inspect the
unmodified profile geometry. In that strict mode `freqTLS`:

* emits a warning that the parameter is **weakly identified** ("consider `bayesTLS` or a
  bootstrap"),
* returns `NA` on the open side (never a fabricated bound), and
* sets a `conf.status` marker (`open_lower`, `open_upper`, or `open_both`).

Here is a deliberately sparse design — few temperatures and little mortality contrast —
that does not identify `CTmax`. We fit it live, then show the user-facing bootstrap
recovery recipe separately. A 1,000-refit bootstrap is intentionally not executed
during package checks; run that displayed chunk interactively when you need the
fallback interval.

```{r non-closing}
set.seed(7)
sparse <- simulate_tls(
  temps  = c(35, 36),     # only two temperatures
  times  = c(1, 2),       # only two durations
  reps   = 2, n = 8,
  CTmax  = 36, z = 4,
  family = "binomial", seed = 7
)
sparse_fit <- suppressWarnings(
  fit_tls(sparse, y = survived, n = total, time = duration, temp = temp,
          family = "binomial", tref = 1)
)
```

```{r bootstrap-fallback-recipe, eval = FALSE}
ci_default <- tryCatch(
  withCallingHandlers(
    confint(sparse_fit, "CTmax", method = "profile",
            fallback = TRUE, nboot = 1000, boot_seed = 7),
    warning = function(w) {
      message("caught warning: ", conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  ),
  error = function(e) e
)
ci_default[, c("parameter", "conf.low", "conf.high", "estimate", "method",
               "conf.status")]
```

Now disable fallback to inspect the strict profile. The open side remains `NA`, with an
`open_*` status, rather than becoming a confident but unsupported bound.

```{r strict-non-closing}
ci_strict <- tryCatch(
  withCallingHandlers(
    confint(sparse_fit, "CTmax", method = "profile", fallback = FALSE),
    warning = function(w) {
      message("caught warning: ", conditionMessage(w))
      invokeRestart("muffleWarning")
    }
  ),
  error = function(e) e
)
ci_strict[, c("parameter", "conf.low", "conf.high", "estimate", "method",
              "conf.status")]
```

The Confidence Eye follows the same contract. With its default `fallback = TRUE`, a
successful bootstrap recovery draws a bootstrap lens. For the strict diagnostic below,
`fallback = FALSE` preserves the open profile and draws a **hollow point with no lens**.

```{r non-closing-eye, fig.alt = "Confidence Eye for a weakly identified fit: an enlarged hollow point estimate with no confidence lens and a direct open-profile annotation, signalling that no finite confidence interval is available."}
# fallback = FALSE so the eye refuses to draw a lens when the profile does not
# close (a hollow point only), matching the honest non-closing contract above.
suppressWarnings(plot_confidence_eye(sparse_fit, parm = "CTmax", method = "profile",
                                     fallback = FALSE))
```

## Calibration: how well do the intervals cover?

A confidence interval is only as good as its coverage. The repository-only
coverage study simulates 200 datasets at a known `CTmax`/`z` under each family,
fits every one by ML, builds 95% profile intervals, and records the empirical
(frequentist) coverage. Its summary is shipped with the package:

```{r coverage}
cov_path <- system.file("extdata", "coverage_results.rds", package = "freqTLS")
if (nzchar(cov_path)) {
  cov <- readRDS(cov_path)
  knitr::kable(
    cov$coverage, digits = 3,
    caption = sprintf("Empirical coverage of 95%% profile CIs (nsim = %d; nominal 0.95).",
                      cov$meta$nsim)
  )
}
```

For the **binomial** model the profile intervals are well calibrated — `CTmax` coverage
sits at the nominal 0.95 and every profile closed. For the **beta-binomial** model the
intervals **under-cover** (`CTmax` ≈ 0.84, `z` ≈ 0.87) at this sample size: estimating the
extra overdispersion parameter makes the likelihood-ratio interval optimistic, and a small
fraction of profiles do not close. The honest reading is to treat beta-binomial profile
intervals as approximate and, when coverage matters, calibrate them with a parametric
bootstrap or use `bayesTLS`. This is exactly the regime the ship stance below warns about.
Repository maintainers can reproduce the study from a source checkout with the
repository-only `data-raw/coverage-study.R` script; that script is not installed with
the package.

## Ship stance

The profile gives fast, prior-free, asymmetry-respecting confidence intervals **when
the MLE is interior and the data identify the target**. For boundary asymptotes, very
sparse designs, overdispersion concentrated at zero, or weak variance components, use
the interval route appropriate to the fitted model. Fixed-effect coordinates in a
random-intercept fit can be profiled under the Laplace approximation; variance
components use log-scale Wald intervals unless you request the random-effects-aware
parametric bootstrap. `freqTLS` warns you when the requested route is weak or
unavailable. It
never claims the profile is universally superior to the Bayesian path. See
`vignette("comparing-to-bayesTLS")` for the side-by-side comparison.
