---
title: "Random effects: hierarchical thermal tolerance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random effects: hierarchical thermal tolerance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.6,
  dpi = 96
)
```

This vignette is for a thermal biologist whose survival assays are **grouped** —
colonies, clutches, populations, tanks, broods — and who wants to model
between-group variation without spending a fixed coefficient on every group. That
is exactly what a **random intercept** does: it treats the per-group deviations as
draws from a Gaussian, estimates a single between-group standard deviation, and
**shrinks** each group's estimate toward the overall mean (partial pooling). In
`freqTLS` a random intercept is available on the thermal-tolerance midpoint
`CTmax`, the thermal-sensitivity `log_z`, and the shape coordinates `low` and
`log_k`.

```{r setup}
library(freqTLS)
```

## Fixed groups vs. a random intercept

With a handful of well-sampled groups you can give each its own **fixed** `CTmax`
(`CTmax ~ group`, see `vignette("getting-started")`). With many groups — or groups
you regard as a sample from a population — a **random intercept**
`CTmax ~ 1 + (1 | group)` is the better tool: it costs one variance parameter
instead of one coefficient per group, and the group estimates borrow strength from
each other. The deviations `b_g ~ N(0, sigma_CTmax^2)` are integrated out by
`TMB`'s Laplace approximation; `sigma_CTmax` is estimated by marginal maximum
likelihood. The no-random-effects path is byte-identical to the fixed-effects
model, so adding `(1 | group)` changes nothing when there is nothing to pool.

## A random intercept on CTmax

We simulate 14 colonies whose CTmax varies with a between-colony SD of 1.5 °C, then
fit a single random intercept.

```{r ctmax-re}
set.seed(1)
d <- simulate_tls(family = "binomial", CTmax = 36, z = 4,
                  re_sd = 1.5, n_re_groups = 14, seed = 1)
fit <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         CTmax ~ 1 + (1 | colony)),
  data = d, family = "binomial", tref = 1, quiet = TRUE)

# The estimates table carries the population CTmax / z and the between-colony SD.
est <- fit$estimates
est[est$parameter %in% c("CTmax", "z", "sigma_CTmax"),
    c("parameter", "estimate", "std.error")]
```

`sigma_CTmax` is the estimated between-colony standard deviation of `CTmax`, in °C
— directly interpretable on the thermal-tolerance scale. It is a maximum-likelihood
variance component, so it is biased **low** when there are few groups: it is
essentially unbiased by ~14 groups but increasingly shrunk toward zero below that.
A focused recovery study (`data-raw/re-recovery-study.R`) makes the bias concrete,
and shows the fixed-effect `CTmax` interval losing coverage in step — the empirical
basis for the advisory `fit_tls()` emits below ~8 groups:

```{r re-recovery, echo = FALSE}
rr_path <- system.file("extdata", "re_recovery_results.rds", package = "freqTLS")
if (nzchar(rr_path) && file.exists(rr_path)) {
  rr <- readRDS(rr_path)
  knitr::kable(
    rr$re_recovery[, c("n_groups", "mean_sigma", "rel_bias", "CTmax_wald_coverage")],
    col.names = c("n groups", "mean sigma_CTmax", "rel. bias", "CTmax 95% coverage"),
    caption = sprintf(
      paste("Random-intercept recovery vs number of groups (true sigma_CTmax = %.1f;",
            "%d sims/cell). The ML SD is biased low and the fixed-effect CTmax",
            "interval under-covers with few groups; both settle by ~14 groups."),
      rr$meta$truth_sigma, rr$meta$nsim),
    digits = 3)
}
```

So with only a handful of groups the reported `sigma_CTmax` understates the true
spread and the `CTmax` interval is optimistic; prefer `confint(method =
"bootstrap")` (or `bayesTLS`) for the fixed effects, and read the variance
component as a lower bound.

The conditional modes (BLUPs) — each colony's shrunken deviation from the
population `CTmax` — come from `ranef()`:

```{r ranef}
head(ranef(fit), 4)
```

`confint()` gives a positive (log-scale) Wald interval for `sigma_CTmax`, and a
genuine profile-likelihood interval for the population `CTmax` under the random
effect (each grid point re-runs the Laplace approximation, so it is slower than a
fixed-effects profile):

```{r ctmax-ci}
suppressMessages(confint(fit, "sigma_CTmax", method = "wald"))[
  , c("parameter", "conf.low", "conf.high")]
suppressMessages(confint(fit, "CTmax", method = "profile", npoints = 8))[
  , c("parameter", "conf.low", "conf.high", "method")]
```

The Confidence Eye stays on Wald intervals for a random-effects fit (a profile
eye would re-run the Laplace at every grid point for every row), which is fast and
honest — it is still a confidence interval, never a posterior.

```{r eye, fig.alt = "Confidence Eye for the population CTmax of a random-intercept fit: a pale confidence lens with a hollow point estimate near 36 degrees Celsius, drawn with a Wald interval; a freqTLS uncertainty display, not a posterior density."}
suppressMessages(plot_confidence_eye(fit, parm = "CTmax"))
```

## Random effects on sensitivity and curve shape

The same `(1 | group)` syntax works on `log_z` (thermal sensitivity) and on the
shape coordinates `low` (the lower, long-exposure asymptote) and `log_k` (the
steepness). One nuance matters: because `z`, `low`, and `k` are modelled on
transformed scales, their variance components live on those scales —
`sigma_logz` is a SD on `log(z)`, `sigma_low` on `logit(low)`, `sigma_logk` on
`log(k)`. Read `sigma_logz` as an approximate *multiplicative* spread on `z`:
`exp(sigma_logz)` is the fold-change across groups, not a z-unit SD.

```{r logz-re}
set.seed(2)
dz <- simulate_tls(family = "binomial", CTmax = 36, z = 4,
                   re_sd_z = 0.3, n_re_groups = 14, seed = 2)
fit_z <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         log_z ~ 1 + (1 | colony)),
  data = dz, family = "binomial", tref = 1, quiet = TRUE)
sg <- fit_z$estimates$estimate[fit_z$estimates$parameter == "sigma_logz"]
c(sigma_logz = round(sg, 3),
  approx_fold_spread_in_z = round(exp(sg), 3))
```

The upper asymptote `up` is the **one shape with no random effect**: it is the
nested gap `up = low + (1 - low) * plogis(...)`, which has no single internal
coordinate (the same reason `up` has no profile-likelihood coordinate). Put the
random intercept on `low` or `log_k` instead.

## Combining random effects, and what is out of scope

You can place random intercepts on more than one coordinate. When two or more
share the **same** grouping factor, `freqTLS` fits them as **independent**
variances (all cross-correlations forced to zero) and warns:

```{r combined}
set.seed(3)
db <- simulate_tls(family = "binomial", CTmax = 36, z = 4,
                   re_sd = 1.2, re_sd_z = 0.3, n_re_groups = 14, seed = 3)
# Both random intercepts on the same `colony` grouping. This fits, but emits a
# warning that the two variances are independent (no correlation term); we pass
# `quiet = TRUE` to silence it here and discuss it below.
fit_both <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         CTmax ~ 1 + (1 | colony), log_z ~ 1 + (1 | colony)),
  data = db, family = "binomial", tref = 1, quiet = TRUE)
fit_both$estimates$parameter[grepl("^sigma", fit_both$estimates$parameter)]
```

In real data the group-level `CTmax` and `z` deviations are usually correlated; by
forcing that correlation to zero, the independent-variance fit absorbs it into the
marginal SDs and the fixed-effect intervals. If you need a **correlated** random
effect — or crossed/nested grouping factors, or a random *slope* — use
[`bayesTLS`](https://github.com/daniel1noble/bayesTLS), whose Stan back end fits
the full covariance. `freqTLS` deliberately stops at independent random
intercepts on `CTmax` / `log_z` / `low` / `log_k`.

## The likelihood view of shrinkage

A Gaussian random effect *is* a prior on the group deviations, and the BLUPs are
shrunk toward zero exactly as a posterior mean would be — the empirical-Bayes view
of mixed models. What stays prior-free is everything else: the fixed effects and
the variance components are maximum-likelihood estimates, and their intervals are
profile-likelihood or Wald confidence intervals, not credible intervals. That
is the same complementary stance the rest of `freqTLS` takes toward `bayesTLS`
(see `vignette("frequentist-and-bayesian")`): the random effect buys you partial
pooling without committing to priors on the parameters you report.
```

