---
title: "Random effects: hierarchical thermal tolerance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random effects: hierarchical thermal tolerance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.6,
  dpi = 96
)
```

This vignette is for a thermal biologist whose survival assays are **grouped** —
colonies, clutches, populations, tanks, broods — and who wants to model
between-group variation without spending a fixed coefficient on every group. That
is exactly what a **random intercept** does: it treats the per-group deviations as
draws from a Gaussian, estimates a single between-group standard deviation, and
**shrinks** each group's estimate toward the overall mean (partial pooling). In
`freqTLS` a random intercept is available on the thermal-tolerance midpoint
`CTmax`, the thermal-sensitivity `log_z`, and the shape coordinates `low` and
`log_k`.

```{r setup}
library(freqTLS)
```

## Fixed groups vs. a random intercept

With a handful of well-sampled groups you can give each its own **fixed** `CTmax`
(`CTmax ~ group`, see `vignette("freqTLS")`). With many groups — or groups
you regard as a sample from a population — a **random intercept**
`CTmax ~ 1 + (1 | group)` is the better tool: it costs one variance parameter
instead of one coefficient per group, and the group estimates borrow strength from
each other. The deviations `b_g ~ N(0, sigma_CTmax^2)` are integrated out by
`TMB`'s Laplace approximation; `sigma_CTmax` is estimated by marginal maximum
likelihood. A formula without `(1 | group)` uses the ordinary fixed-effects
objective. Adding `(1 | group)` is not a no-op: it adds and integrates a
group-level Gaussian block, even when its fitted variance is near zero.

## A random intercept on CTmax

We simulate 14 colonies whose CTmax varies with a between-colony SD of 1.5 °C, then
fit a single random intercept.

```{r ctmax-re}
set.seed(1)
d <- simulate_tls(family = "binomial", CTmax = 36, z = 4,
                  re_sd = 1.5, n_re_groups = 14, seed = 1)
fit <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         CTmax ~ 1 + (1 | colony)),
  data = d, family = "binomial", tref = 60, quiet = TRUE)

# The estimates table carries the population CTmax / z and the between-colony SD.
est <- fit$estimates
est[est$parameter %in% c("CTmax", "z", "sigma_CTmax"),
    c("parameter", "estimate", "std.error")]
```

`sigma_CTmax` is the estimated between-colony standard deviation of `CTmax`, in °C
— directly interpretable on the thermal-tolerance scale. It is a maximum-likelihood
variance component, so it is biased **low** when there are few groups: it is
essentially unbiased by ~14 groups but increasingly shrunk toward zero below that.
A focused recovery study makes the bias concrete and shows the fixed-effect
`CTmax` interval losing coverage in step — the empirical basis for the advisory
`fit_tls()` emits below ~8 groups. Repository maintainers can regenerate it with
the source-only `data-raw/re-recovery-study.R` script, which is not installed:

```{r re-recovery, echo = FALSE}
rr_path <- system.file("extdata", "re_recovery_results.rds", package = "freqTLS")
if (nzchar(rr_path) && file.exists(rr_path)) {
  rr <- readRDS(rr_path)
  knitr::kable(
    rr$re_recovery[, c("n_groups", "mean_sigma", "rel_bias", "CTmax_wald_coverage")],
    col.names = c("n groups", "mean sigma_CTmax", "rel. bias", "CTmax 95% coverage"),
    caption = sprintf(
      paste("Random-intercept recovery vs number of groups (true sigma_CTmax = %.1f;",
            "%d sims/cell). The ML SD is biased low and the fixed-effect CTmax",
            "interval under-covers with few groups; both settle by ~14 groups."),
      rr$meta$truth_sigma, rr$meta$nsim),
    digits = 3)
}
```

So with only a handful of groups the reported `sigma_CTmax` understates the true
spread and the `CTmax` interval is optimistic. Prefer `confint(method =
"bootstrap")` (or `bayesTLS`) for the fixed effects, and describe the variance
component as likely downward-biased rather than treating it as a precise estimate
or a literal confidence bound.

The conditional modes (BLUPs) — each colony's shrunken deviation from the
population `CTmax` — come from `ranef()`:

```{r ranef}
head(ranef(fit), 4)
```

Prediction distinguishes the population curve from an observed colony's
conditional curve. Population predictions set the random intercept to zero;
conditional predictions add the fitted BLUP and require the grouping column in
`newdata`:

```{r re-predict}
colony_1 <- as.character(ranef(fit)$group[1])
new_colony <- data.frame(temp = 36, duration = 2, colony = colony_1)
data.frame(
  target = c("population", "colony"),
  survival = c(
    predict(fit, new_colony, re.form = "population"),
    predict(fit, new_colony, re.form = "conditional")
  )
)
```

Calling `predict()` without `re.form` on a random-effects fit warns and returns
the population prediction. An unseen group cannot receive an estimated BLUP, so
conditional prediction stops and points to `re.form = "population"`.

`confint()` routes uncertainty by target. It gives a positive, log-scale Wald
interval for `sigma_CTmax`. It profiles the population `CTmax` under the random
effect, re-running the Laplace approximation at each grid point, so this is slower
than a fixed-effects profile. If that fixed-effect profile does not close, the
default fallback is Wald; request `method = "bootstrap"` to redraw the random
effects and refit the full model instead.

```{r ctmax-ci}
suppressMessages(confint(fit, "sigma_CTmax", method = "wald"))[
  , c("parameter", "conf.low", "conf.high")]
suppressMessages(confint(fit, "CTmax", method = "profile", npoints = 8))[
  , c("parameter", "conf.low", "conf.high", "method")]
```

The Confidence Eye always uses Wald intervals for a random-effects fit, even when
`method = "profile"` is requested, because a profile eye would re-run the Laplace
approximation at every grid point for every row. Use `confint(fit, method =
"profile")` for fixed-effect profiles or `confint(fit, method = "bootstrap")`
for random-effects-aware bootstrap intervals. The eye remains a confidence-interval
display, never a posterior.

```{r eye, fig.alt = "Confidence Eye for the population CTmax of a random-intercept fit: a pale outlined confidence lens with a dark centre mark and hollow estimate near 36 degrees Celsius, drawn with a Wald interval; a freqTLS uncertainty display, not a posterior density."}
suppressMessages(plot_confidence_eye(fit, parm = "CTmax"))
```

## Random effects on sensitivity and curve shape

The same `(1 | group)` syntax works on `log_z` (thermal sensitivity) and on the
shape coordinates `low` (the lower, long-exposure asymptote) and `log_k` (the
steepness). One nuance matters: because `z`, `low`, and `k` are modelled on
transformed scales, their variance components live on those scales —
`sigma_logz` is a SD on `log(z)`, `sigma_low` on `logit(low)`, `sigma_logk` on
`log(k)`. Read `sigma_logz` as an approximate *multiplicative* spread on `z`:
`exp(sigma_logz)` is the fold-change across groups, not a z-unit SD.

```{r logz-re}
set.seed(2)
dz <- simulate_tls(family = "binomial", CTmax = 36, z = 4,
                   re_sd_z = 0.3, n_re_groups = 14, seed = 2)
fit_z <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         log_z ~ 1 + (1 | colony)),
  data = dz, family = "binomial", tref = 60, quiet = TRUE)
sg <- fit_z$estimates$estimate[fit_z$estimates$parameter == "sigma_logz"]
c(sigma_logz = round(sg, 3),
  approx_fold_spread_in_z = round(exp(sg), 3))
```

The upper asymptote `up` is the **one shape with no random effect**: under the
disjoint-bounds parameterisation `up = up_min + up_w * plogis(beta_up)` it has its
own coordinate `beta_up`, but the compiled objective carries no random-intercept
term for `up` (and its profile path is not wired either — the same gap that leaves
`up` with a delta-method interval). Put the random intercept on `low` or `log_k`
instead.

## Combining random effects, and what is out of scope

You can place random intercepts on more than one coordinate. When two or more
share the **same** grouping factor, `freqTLS` fits them as **independent**
variances (all cross-correlations forced to zero) and warns:

```{r combined}
set.seed(3)
db <- simulate_tls(family = "binomial", CTmax = 36, z = 4,
                   re_sd = 1.2, re_sd_z = 0.3, n_re_groups = 14, seed = 3)
# Both random intercepts on the same `colony` grouping. This fits, but emits a
# warning that the two variances are independent (no correlation term); we pass
# `quiet = TRUE` to silence it here and discuss it below.
fit_both <- fit_tls(
  tls_bf(survived | trials(total) ~ time(duration) + temp(temp),
         CTmax ~ 1 + (1 | colony), log_z ~ 1 + (1 | colony)),
  data = db, family = "binomial", tref = 60, quiet = TRUE)
fit_both$estimates$parameter[grepl("^sigma", fit_both$estimates$parameter)]
```

In real data the group-level `CTmax` and `z` deviations are usually correlated; by
forcing that correlation to zero, the independent-variance fit absorbs it into the
marginal SDs and the fixed-effect intervals. If you need a **correlated** random
effect — or crossed/nested grouping factors, or a random *slope* — use
[`bayesTLS`](https://github.com/daniel1noble/bayesTLS), whose Stan back end fits
the full covariance. `freqTLS` deliberately stops at independent random
intercepts on `CTmax` / `log_z` / `low` / `log_k`.

## The likelihood view of shrinkage

A Gaussian random effect is a distributional model for the latent group deviations,
not a prior placed on the reported population parameters. Its conditional modes
(BLUPs) have an empirical-Bayes interpretation and shrink toward zero. The fixed
effects and variance components are estimated by marginal maximum likelihood, and
their intervals are
profile-likelihood intervals for fixed-effect coordinates, log-scale Wald intervals
for variance components, or parametric-bootstrap intervals when explicitly requested;
they are not credible intervals. That
is the same complementary stance the rest of `freqTLS` takes toward `bayesTLS`
(see `vignette("frequentist-and-bayesian")`): the random effect buys you partial
pooling without committing to priors on the parameters you report.
